---
title: "Pedalling Forward: The Evolution of Dedicated Cycling Infrastructure in Canadian Cities from 2010 to 2022"
subtitle: "R Code for Figures and Tables"
author:
- "Richard Wen richard.wen@utoronto.ca"
- "Konrad Samsel konrad.samsel@mail.utoronto.ca"
- "Brice Batomen brice.kuimi@utoronto.ca"
- "Linda Rothman linda.rothman@torontomu.ca"
date: "`r format(Sys.time(), '%B %d, %Y')`"
knit: |
    (function(input_rmd, ...) {
    rmarkdown::render(
        input_rmd,
        rmarkdown::html_document(
            toc = TRUE,
            toc_float = TRUE,
            highlight = "zenburn",
            code_folding = "hide",
            df_print = "paged",
            self_contained = FALSE
        ),
        output_dir = "../docs",
        output_file = "index", ...)
    })
---

```{r include = FALSE, echo = FALSE}
knitr::opts_chunk$set(
	warning = FALSE,
	message = FALSE
)
```

# Installation

1. Install [R](https://www.r-project.org/)
2. Install [RTools](https://cran.r-project.org/) if you are on Windows
3. Install [RStudio](https://posit.co/download/rstudio-desktop/)

For more details, see [Software and Package Versions](#software-and-package-versions).

# Running This Code

1. Ensure the installation steps above are completed
2. Download a zip of the code and data [here](https://github.com/rrwen/recovr-infracycle/archive/refs/heads/main.zip) and unzip it
	* Code Repository: [github.com/rrwen/recovr-infracycle](https://github.com/rrwen/recovr-infracycle)
3. In RStudio, open the [src/src.Rproj](https://github.com/rrwen/recovr-infracycle/blob/main/src/src.Rproj) file
4. Then, open the [src/index.Rmd](https://github.com/rrwen/recovr-infracycle/blob/main/src/index.Rmd) file
6. In RStudio:
	* Run all code: Click the `Run` drop down (top right of the code pane) and click `Run All`
	* Generate HTML version: Click `knit` (top left of code pane) and a file will be generated in `docs/index.html` (you may have to try a couple times to succeed due to `webshot2` timeouts)

# Libraries

Install R packages if needed.

```{r warning = FALSE, message = FALSE, results = FALSE}

# Required packages
required_packages <- c(
	"rmarkdown",
	"bookdown",
	"knitr",
	"tidyverse",
	"glue",
	"readxl",
	"ggtext",
	"scales",
	"patchwork",
	"DiagrammeR",
	"DiagrammeRsvg",
	"webshot2",
	"magick",
	"rsvg",
	"sf",
	"tmap",
	"ggspatial",
	"prettymapr",
	"units",
	"lubridate",
	"kableExtra"
)

# Try to install packages if not installed
default_options <- options()
tryCatch(
	{
		# Disable interactivity
		options(install.packages.compile.from.source = "always")
		
		# Install package if not installed
		for (package in required_packages) {
			is_package_installed <- require(package, character.only = TRUE)
			if (!is_package_installed & package != "osmplotr") {
				cat(paste0("Installing package: ", package, "\n"))
				install.packages(package)
			} else {
				cat(paste0("Package already installed: ", package, "\n"))
			}
		}
	},
	error = function(cond) {
		stop(cond)
	},
	finally = {
		options(default_options) # reset interactivity
	}
)
```

Load R libraries.

```{r warning = FALSE, message = FALSE}
library(DiagrammeR)
library(DiagrammeRsvg)
library(ggtext)
library(glue)
library(kableExtra)
library(lubridate)
library(patchwork)
library(readxl)
library(rsvg)
library(sf)
library(tidyverse)
library(tmap)
library(ggspatial)
library(units)
```

# Settings

```{r}
settings <- list()

# Infrastructure types in order
settings$type_recode_infra <- c(
	PBL = "Cycle Track",
	BUF = "Buffered Lane",
	PL = "Painted Lane",
	LSB = "Local Street\nBikeway"
)

# Infrastructure types to remove
settings$type_filter_infra <- c("N", "None", "SR")

# Road types in order
settings$type_recode_road <- c(
	Arterial = "Arterial",
	Collector = "Collector",
	Local = "Local"
)

# Column references
settings$year_col_road <- "verify_install_year"
settings$type_col_road <- "road_type_recode"
settings$type_col_infra <- "verify_install_type"

# Set years of interest
settings$year_min <- 2009
settings$year_max <- 2022

# Plot settings
settings$line_year <- 2019
settings$basemaps <- c(
	"CartoDB.Positron",
	"CartoDB.DarkMatter",
	"Esri.WorldGrayCanvas"
)

# Map infrastructure changes since year
settings$infra_changes_year <- 2020

# Apply map settings
tmap_options(basemaps = settings$basemaps)
```

# Functions

## Function 1: calc_yearly_len

Calculate yearly road lengths.

The following function calculates yearly road lengths by infrastructure type using cumulative sums and filling in missing years and types.

For a given infrastructure type, the total road length for a given year is expressed below:

$$
length_{year,type} = f(year,type) = \sum_{i=year_{min}}^{year}l_{i, type}\ \mid\ l_{i, type} \geq 0
$$

Where:

-   $year$ is the given year
-   $type$ is the infrastructure type
-   $year_{min}$ is the earliest year available in the data
-   $l_{i,type}$ is the road length $l$ for previous years $i$ and infrastructure $j$
-   $l_{i,type}$ is set to 0 if there is no data

```{r}

#' Calculate Yearly Road Lengths By Infrastructure Type
#' 
#' Calculates the cumulative yearly road lengths by infrastructure type without considering infrastructure changes.
#'
#' @param df A data.frame with three columns containing the year, type, and road lengths.
#' @param year_col The name (char) or index (int) of the column containing the years.
#' @param type_col The name (char) or index (int) of the column containing the infrastructure type
#' @param len_col The name (char) or index (int) of the column containing the road lengths.
#' @param out_col The name (char) of the column containing the calculated yearly road lengths by type.
#'
#' @return A data.frame with three columns containing the year, type, and calculated yearly road lengths by type.
#' @export
#'
calc_yearly_len <- function(
		df,
		year_col = "verify_install_year",
		type_col = "verify_install_type",
		len_col = "geometry_len_km",
		out_col = "len",
		year_min = settings$year_min,
		year_max = settings$year_max
	) {
	
	# Convert data types
	df <- as.data.frame(df)
	df[[year_col]] <- as.integer(df[[year_col]])
	df[[type_col]] <- as.character(df[[type_col]])
	df[[len_col]] <- as.numeric(df[[len_col]])
	
	# Remove rows with empty type
	out <- df %>% filter(
		!is.na(.data[[type_col]])
	)
	
	# Filter to min and max years
	if (year_min > 0) {
		df <- df %>% filter(
			.data[[year_col]] >= year_min
		)
	} else {
		year_min <- min(out[[year_col]], na.rm = TRUE)
	}
	if (year_max > 0) {
		df <- df %>% filter(
			.data[[year_col]] <= year_max
		)
	} else {
		year_max <- max(out[[year_col]], na.rm = TRUE)
	}
	
	# Add dummy len for each type and year combo
	# Covers cases where type and year combo does not exist
	# E.g. No new PL installs in 2021, hence a record PL in 2021 does not exist
	type_uniq <- unique(out[[type_col]])
	type_n <- length(type_uniq)
	year_uniq <- year_min:year_max
	year_n <- length(year_uniq)
	out <- out %>% add_row(
		!!year_col := rep(year_uniq, each = type_n),
		!!type_col := rep(type_uniq, year_n),
		!!len_col := rep(0, type_n * year_n)
	)
	
	# Calc cumsum for each non-empty type ordered by year
	out <- out %>%
		arrange(.data[[year_col]]) %>%
		group_by(.data[[type_col]]) %>%
		mutate(
			!!out_col := cumsum(.data[[len_col]])
		)

	# Get the last cumsum for each year and type
	out <- out %>%
		group_by(.data[[year_col]], .data[[type_col]]) %>%
		arrange(desc(row_number())) %>%
		slice(1)
	
	# Return only the columns spec
	out <- out %>% select(c(
			year_col,
			type_col,
			out_col
		))
	return(out)
}
```

## Function 2: calc_yearly_adj_len

Calculate yearly adjusted road length.

The following function calculates yearly adjusted road lengths by infrastructure type using cumulative sums and filling in missing years and types.

For a given infrastructure type, the total adjusted road length for a given year is expressed below:

$$
length_{year,type}^{install} + length_{year,type}^{change_i} - length_{year,type}^{replacement_i}
$$ Where:

-   $length_{year,type}^{install}$ are the yearly cumulative road lengths for an infrastructure $type$ installation
-   $length_{year,type}^{change_i}$ are the yearly cumulative road lengths for an infrastructure $type$ change in order $i$
-   $length_{year,type}^{replacement_i}$ are the yearly cumulative road lengths for an infrastructure $type$ replaced by change in order $i$

```{r}

#' Calculate Yearly Adjusted Road Lengths By Infrastructure Type
#' 
#' Calculates the cumulative yearly adjusted road lengths by infrastructure type accounting for installations and subsequent changes.
#'
#' @param df A data.frame with three columns containing the year, type, and road lengths.
#' @param year_cols A vector of the names (char) or indices (int) of the columns containing the years of installations followed by infrastructure changes in order.
#' @param type_cols A vector of the names (char) or indices (int) of the columns containing the infrastructure types of installations followed by infrastructure changes in order.
#' @param type_col The name (char) of the column containing the type.
#' @param len_cols A vector of the names (char) or indices (int) of the columns containing the road lengths of installations followed by infrastructure changes in order.
#' @param out_cols The name (char) of the column containing the calculated yearly road lengths by type.
#' @param out_col The name (char) of the column containing the calculated yearly adjusted road lengths by type.
#' @param repl_suffix A suffix (char) to append to the columns representing the road lengths of replaced infrastructure types from changes.
#' @param ... Additional arguments passed to calc_yearly_len.
#' 
#' @return A data.frame with columns containing the year, type, cumulative road lengths of installations, changes, and replacements, and calculated yearly adjusted road lengths by type.
#' @export
#'
calc_yearly_adj_len <- function(
		df,
		year_cols = c("verify_install_year", "verify_upgrade1_year", "verify_upgrade2_year"),
		type_cols = c("verify_install_type", "verify_upgrade1_type", "verify_upgrade2_type"),
		type_col = "type",
		len_cols = "geometry_len_km",
		out_cols = c("install_len", "upgrade1_len", "upgrade2_len"),
		out_col = "adj_len",
		repl_suffix = "_replaced",
		...
	) {
	
	# Ensure df
	df <- as.data.frame(df)
	
	# Convert len_col if char
	len_cols <- rep(len_cols, length(year_cols))
	
	# Check cols same size
	year_cols_n <- length(year_cols)
	type_cols_n <- length(type_cols)
	len_cols_n <- length(len_cols)
	out_cols_n <- length(out_cols)
	if (length(unique(c(year_cols_n, type_cols_n, len_cols_n, out_cols_n))) != 1) {
		stop(glue(
			"The arguments 'year_cols' ({year_cols_n}), 'type_cols' ({type_cols_n}), 'len_cols' ({len_cols_n}), and 'out_cols' ({out_cols_n}) must be the same length."
		))
	}
	
	# Calc yearly lens by infra type per install or change
	out <- list()
	for (i in 1:length(year_cols)) {
		
		# Get year, type, and len cols
		ycol <- year_cols[[i]]
		tcol <- type_cols[[i]]
		lcol <- len_cols[[i]]
		ocol <- out_cols[[i]]
		
		# Calc yearly len for install or change
		has_infra <- !is.na(df[[tcol]]) %>% all
		if (has_infra) {
			out <- append(
				out,
				calc_yearly_len(
					df,
					year_col = ycol,
					type_col = tcol,
					len_col = lcol,
					out_col = ocol,
					...
				) %>%
					rename(
						"year" := !!ycol,
						"type" := !!tcol
					) %>% list
			)
		}
		
		# Calc yearly len for replacement
		if (i > 1) {
			
			# Get repl cols
			tcol_repl <- type_cols[[i - 1]]
			lcol_repl <- len_cols[[i - 1]]
			
			# Filter for repl records only where type is not eq to change type
			df_repl <- df %>% filter(.data[[tcol]] != .data[[tcol_repl]])
			
			# Calc repl len if there are any changes
			has_change <- !is.na(df_repl[[tcol]]) %>% all 
			if (has_change) {
				out <- append(
					out,
					calc_yearly_len(
						df_repl,
						year_col = ycol,
						type_col = tcol_repl,
						len_col = lcol_repl,
						out_col = glue("{ocol}{repl_suffix}"),
						...
					) %>%
					rename(
						"year" := !!ycol,
						"type" := !!tcol_repl
					) %>% list
				)
			}
		}
	}
	
	# Combine all lens in list to single df
	out <- out %>%
		reduce(
			left_join, by = c("year", "type")
		) %>%
		ungroup()
	
	# Create template for change and repl cols
	change_cols <- paste0(out_cols[2:out_cols_n])# change cols
	change_cols <- c(change_cols, paste0(out_cols[2:out_cols_n], repl_suffix)) # repl cols
	change_cols_add <- rep(0, length(change_cols)) # set default vals
	names(change_cols_add) <- change_cols
	
	# Add change and repl cols set to 0 if not present
	out <- out %>% add_column(
		!!!change_cols_add[setdiff(names(change_cols_add), names(.))]
	)
	
	# Set NA to 0
	out <- out %>% mutate(
		across(everything(), ~replace_na(., 0))
	)
	
	# Calc yearly adj lens by infra type
	out <- out %>%
		mutate( # added len by infra types due to install or changes
			!!out_col := reduce(across(all_of(out_cols)), `+`)
		) %>%
		mutate( # removed len by infra types due to replacements
			!!out_col := .data[[out_col]] - reduce(
				across(all_of(
					paste0(out_cols[2:out_cols_n], repl_suffix)
				)),
				`-`
			)
		)
	
	# Rename type col
	out <- out %>% rename(!!type_col := type)
	return(out)
}
```

## Function 3: plot_yearly_len

Plot road lengths by generic types.

This function plots an area chart showing the cumulative road lengths by a user-defined type for each year.

This is a generic function for user-defined types such as infrastructure or road types.

```{r}

#' Plot Yearly Road Lengths By Type
#' 
#' Creates an area plot of road lengths by category types.
#'
#' @param df A data.frame with three columns containing the year, type, and road lengths.
#' @param title The title (char) of the plot.
#' @param title_underline Set to TRUE to underline the title.
#' @param x_title The title (char) of the x-axis.
#' @param y_title The title (char) of the y-axis.
#' @param y_suffix The suffix (char) to add to the end of y axis values.
#' @param y_lim Minimum and maximum road length (numeric) as a vector of length 2 to limit the range of the y-axis. Set to `NULL` for auto.
#' @param legend_title The title (char) of the legend.
#' @param legend Set to TRUE to include a legend.
#' @param year_col The name (char) or index (int) of the column containing the years.
#' @param year_min The minimum year (int) to display.
#' @param year_max The maximum year (int) to display.
#' @param year_int The year intervals (int) to display. For example, 1 displays every year, and 2 displays every two years.
#' @param len_col The name (char) or index (int) of the column containing the road lengths.
#' @param len_per Set to `TRUE` to add final percentages at the ending year or `FALSE` to omit this.
#' @param type_col The name (char) or index (int) of the column containing the type.
#' @param type_filter A vector (char) of types to remove fomr the plot.
#' @param type_recode A named vector (char) of names representing types and values representing the values to replace each type with.
#' @param line_km The km (numeric) of the red reference line.
#' @param line_show Set to TRUE to draw the km red reference line.
#' @param line_year Set to a year (int) to draw a reference line for a year. If FALSE, a line will not be drawn.
#' @param color_low The bottom color (char) of the type.
#' @param color_high The top color (char) of the type.
#' @return An area ggplot of the cumulative yearly road lengths by type.
#' @export
#'
plot_yearly_len <- function(
        df,
        title = "",
        title_underline = TRUE,
        x_title = "",
        y_title = "",
        y_suffix = " km",
        y_lim = NULL,
        legend_title = "Type",
        legend = TRUE,
        year_col = "year",
        year_min = FALSE,
        year_max = FALSE,
        year_int = 1,
        len_col = "adj_len",
        len_per = FALSE,
        type_col = "type",
        type_filter = c(),
        type_recode = c(),
        line_km = 10,
        line_show = FALSE,
        line_year = FALSE,
        color_low = "#DFEBF7",
        color_high = "#3683BB"
) {
	
	# Filter to start and end years
	if (year_min > 0) {
		df <- df %>% filter(
			.data[[year_col]] >= year_min
		)
	}
	if (year_max > 0) {
		df <- df %>% filter(
			.data[[year_col]] <= year_max
		)
	}
	
	# Filter out particular infrastructure types
	if (length(type_filter) > 0) {
		df <- df %>% filter(
			!.data[[type_col]] %in% type_filter
		)
	}
	
	# Recode and reorder category types
	if (length(type_recode) > 0) {
		
		# Reorder category types
		type_uniq <- unique(df[[type_col]])
		type_reorder <- names(type_recode)
		type_reorder <- c(type_reorder, type_uniq[!type_uniq %in% type_reorder])
		df[[type_col]] <- factor(df[[type_col]], levels = type_reorder)
		
		# Recode category types
		df[[type_col]] <- recode(df[[type_col]], !!!type_recode)
	}
	
	# Create fill colors
	type_n <- length(type_uniq)
	type_colors <- scales::seq_gradient_pal(
		color_low,
		color_high
	)(seq(0, 1, length.out = type_n))
	
	# Create base area plot with legend and labels
	len_max <- max(df[[len_col]], na.rm = TRUE)
	year_max <- max(df[[year_col]], na.rm = TRUE)
	out <- ggplot(
		df,
		aes(
			x = .data[[year_col]],
			y = .data[[len_col]],
			fill = .data[[type_col]],
			order = desc(.data[[type_col]])
		)
	) +
	geom_area(colour = NA, alpha = 0.7) +
	scale_fill_manual(values = type_colors) +
	geom_line(
		position = "stack",
		size = 0.2
	) +
	labs(
		x = x_title,
		y = y_title,
		fill = legend_title
	) +
	guides(
		fill = FALSE,
		color = FALSE
	) +
	scale_x_continuous(
		breaks = seq(year_min, year_max, by = year_int),
		labels = seq(year_min, year_max, by = year_int),
		limits = c(year_min, if (len_per) year_max + 1 else year_max)
	) +
	scale_y_continuous(
		label = scales::label_number(suffix = y_suffix)
	) +
	theme_minimal() +
	theme(
		plot.margin = unit(c(5,5,5,5), "points")
	)
	
	# Scale road length axis y
	if (!is.null(y_lim)) {
		out <- out + ylim(y_lim)
	}
	
	# Add title
	if (title_underline) {
		out <- out + ggtitle(
			bquote(underline(.(title)))
		)
	} else {
		out <- out + ggtitle(title)
	}
	
	# Add legend
	if (legend) {
		out <- out + guides(fill = guide_legend(
			reverse = FALSE,
			override.aes = list(
				alpha = 0.7,
				color = NA,
				shape = NA
			)
		))
	}
	
	# Add percentages
	if (len_per) {
		df_perc <-  df %>% filter(
			.data[[year_col]] == year_max
		) %>% arrange(desc(.data[[type_col]])) %>% mutate(
			len = cumsum(.data[[len_col]]) - (.data[[len_col]] / 2),
			perc = .data[[len_col]] / sum(.data[[len_col]], na.rm = T),
			perc = paste0(round(perc * 100, 1), "%")
		)
		out <- out + geom_text(
			data = df_perc,
			x = year_max,
			size = 2.75,
			hjust = -0.225,
			aes(
				y = len,
				label = perc
			)
		)
	}
	
	# Add dotted year ref line
	if (line_year) {
		out <- out + geom_vline(
			xintercept = line_year,
			color = "black",
			linetype = "dashed"
		)
	}
	
	# Add red 50km ref line
	if (line_show) {
		out <- out + geom_segment( # 50km red line
			aes(
				x = 2009,
				y = 0,
				xend = 2009,
				yend = line_km,
				color = "#bb0000"
			)
		) +
		geom_segment( # 50km red triangle point down
			aes(
				x = 2009,
				y = (line_km + 0.01) - (len_max * 0.05),
				xend = 2009,
				yend = line_km - (len_max * 0.05),
				color = "#bb0000"
			),
			arrow = arrow(
				length = unit(0.03, "npc"),
				ends = "last",
				type = "closed"
			)
		) +
		geom_segment( # 50km red triangle point up
			aes(
				x = 2009,
				y = (len_max * 0.05) - 0.01,
				xend = 2009,
				yend = (len_max * 0.05),
				color = "#bb0000"
			),
			arrow = arrow(
				length = unit(0.03, "npc"),
				ends = "last",
				type = "closed"
			)
		) +
		annotate(
			"text",
			x = 2009,
			y = line_km,
			label = paste0(line_km, "km"),
			color = "#bb0000",
			hjust = -0.225
		)
	}
	return(out)
}
```

## Function 3a: plot_yearly_len_infra

Plot yearly adjusted road lengths by infrastructure type.

This function plots area charts of yearly road lengths by infrastructure types for a list of data.

This uses the `plot_yearly_len` function.

```{r}

#' Plot Yearly Road Lengths By Infrastructure Type
#' 
#' Creates area plots of road lengths by infrastructure type.
#'
#' @param df_list A list of lists, where each key is the title and each value contains a list with the following structure:
#' \itemize{
#'   \item \code{data}: data.frame containing the install and change years, type, and road segment lengths.
#'   \item \code{roadway_total}: the total roadway length if `rodway_per` is given. This is used as the denominator to normalize road lengths. 
#'.  \item \code{roadway_per}: Number of units of total roadway length (numeric) to normalize by (e.g. 1000 means per 1000 km of roadway). Set to `NULL` or omit to disable normalization of road lengths.
#' }
#' @param len_title The title (char) for the road lengths.
#' @param line_show Set to `TRUE` to add a 50km reference line.
#' @param ... Additional arguments passed to `plot_yearly_len`.
#'
#' @return Multiple area ggplots of the cumulative yearly road lengths by infrastructure type combined with patchwork.
#' @export
#'
plot_yearly_len_infra <- function(
		df_list,
		len_title = "Total length per 1000 centreline-km of roadway",
		line_km = 10,
		line_show = TRUE,
		...
	) {
	
	# Create infra plots from data
	p <- list()
	for (i in 1:length(df_list)) {
		
		# Get data and plot title
		df <- df_list[[i]]$data
		ptitle <- names(df_list)[[i]]
		
		# Get roadway vars if exists
		roadway_per <- NULL
		roadway_total <- NULL
		if ("roadway_per" %in% names(df_list[[i]])) {
			roadway_per <- df_list[[i]]$roadway_per
		}
		if ("roadway_total" %in% names(df_list[[i]])) {
			roadway_total <- df_list[[i]]$roadway_total
		}
		
		# Calc infra per year
		p[[i]] <- calc_yearly_adj_len(
			df,
			type_col = settings$type_col_infra
		)
		
		# Norm len if needed
		len_col <- "adj_len"
		if (!is.null(roadway_per)) {
			p[[i]] <- p[[i]] %>% mutate(
				adj_len_norm = 
					(adj_len / roadway_total) * roadway_per
			)
			len_col = "adj_len_norm"
		}
		
		# Add infra plot
		p[[i]] <- p[[i]] %>% plot_yearly_len(
		    	title = ptitle,
		    	year_min = settings$year_min,
		        year_max = settings$year_max,
				type_col = settings$type_col_infra,
				type_filter = settings$type_filter_infra,
		        type_recode = settings$type_recode_infra,
				legend_title = "Infrastructure Type",
				line_km = line_km,
				line_show = line_show,
		        line_year = settings$line_year,
				len_col = len_col,
				...
		    )
	}
	
	# Y-axis title
	y_title <- ggplot() +
		annotate(
			geom = "text",
			x = 1,
			y = 1,
			label = len_title,
			angle = 90,
			size = 5
		) +
		coord_cartesian(clip = "off")+
		theme_void()
	
	# Combine all infra plots together
	out <- (y_title | wrap_plots(p, nrow = length(p))) +
		plot_annotation(
			title = "Roadways with Dedicated Cycling Infrastructure",
			caption = sprintf("Years (%s-%s)", settings$year_min, settings$year_max),
			theme = theme(
				plot.title = element_text(hjust = 0.5, size = 16),
				plot.caption = element_text(hjust = 0.5, size = 14)
			)
		) +
		plot_layout(widths = c(0.05, 1))
	return(out)
}
```

## Function 3b: plot_yearly_len_road

Plots yearly adjusted road lengths for road types.

This function plots area charts of yearly road lengths by overall road type and by infrastructure separated by each road type.

This uses the `plot_yearly_len` function.

```{r}

#' Plot Yearly Road Lengths By Road Type
#'
#' Creates area plots of road lengths by overall road type, and by infrastructure per road type.
#'
#' @param df The data.frame containing the install and change years, type, and road segment types and lengths. 
#' @return Multiple area ggplots of the cumulative yearly road lengths by road type combined with patchwork.
#' @export
#'
plot_yearly_len_road <- function(df, title = "Roadways with Dedicated Cycling Infrastructure") {
	
	# Create list to store plots
	p <- list()

	# Plot overall road types
	p[[1]] <- calc_yearly_len(
		df,
		year_col = settings$year_col_road,
		type_col = settings$type_col_road
	) %>%
	    plot_yearly_len(
	    	title = title,
	    	title_underline = FALSE,
	    	year_col = settings$year_col_road,
			year_min = settings$year_min,
	        year_max = settings$year_max,
			x_title = sprintf("Years (%s-%s)", settings$year_min, settings$year_max),
			y_title = "Total Length (Centreline km)",
			legend_title = "Roadway Type",
			type_col = settings$type_col_road,
			type_recode = settings$type_recode_road,
			len_col = "len",
			line_show = FALSE,
			line_year = settings$line_year,
	        color_low = "#C1DDB3",
	        color_high = "#297A22"
	    ) +
		theme(
			plot.title = element_text(size = 18),
			plot.margin = margin(0, 0, 0, 0, "pt")
		)
	
	# Plot arterial, collector, and local road by infra
	rtypes <- c("Arterial", "Collector", "Local")
	for (i in 1:length(rtypes)) {
		
		# Get road type
		r <- rtypes[i]
		
		# Create infra plot for road type
		p[[i + 1]] <- calc_yearly_adj_len(
			df %>% filter(.data[[settings$type_col_road]] == r),
			type_col = settings$type_col_infra
		) %>%
			plot_yearly_len(
			    title = sprintf("%s Roadways", r),
			    title_underline = FALSE,
			    line_show = FALSE,
			    line_year = settings$line_year,
			    year_int = 2,
			    x_title = sprintf("Years (%s-%s)", settings$year_min, settings$year_max),
			    y_title = "Total Length (Centreline km)",
			    year_min = settings$year_min,
		        year_max = settings$year_max,
				type_col = settings$type_col_infra,
				type_filter = settings$type_filter_infra,
		        type_recode = settings$type_recode_infra,
				legend_title = "Infrastructure Type"
			) +
			theme(
				plot.title = element_text(size = 14),
				plot.margin = margin(0, 12, 0, 0, "pt")
			)
	}
	
	# Plot horizontal gradient bar
	grad_bar <-  ggplot(data.frame(x = 1:4), aes(x = x, y = 1, color = x)) +
		geom_line(size = 4) +
		scale_color_gradient(low = "#C1DDB3", high = "#297A22") +
		theme_void() +
		guides(color = FALSE) +
		theme(
			axis.title = element_blank(),
	        axis.text = element_blank(),
	        axis.ticks = element_blank(),
	        axis.line = element_blank(),
			plot.margin = margin(0, 0, 0, 0, "pt")
		)
	
	# Plot overall and road type plots together
	out <- ( # overall plot
		plot_spacer() +
		p[[1]] +
		plot_spacer() +
		plot_layout(
			widths = c(0.25, 0.35, 0.2)
		)
	) / ( # gradient bar
		plot_spacer() +
		grad_bar +
		plot_spacer() +
		plot_layout(widths = c(-0.8, 10, -1.1))
	) / ( # infra plots
		p[[2]] +
		p[[3]] +
		p[[4]]
	) + plot_layout(
		heights = c(12, 1, 8)
	) + plot_annotation( # A B tags
		tag_levels = list(c("A", "", "B", "", ""))
	) & theme(
		plot.tag = element_text(face = "bold", size = 12)
	)
	return(out)
}
```

## Function 4: plot_yearly_diff

Plots differences between two years.

This function plots a bar chart of differences between two columns containing years.

This function is used to check the differences in installation years between the city's data and the verified data.

```{r}
#' Plot Yearly Differences
#'
#' Creates a bar plot of the differences between two years.
#'
#' @param df The data.frame containing the two columns with the years.
#' @param year_col1 The name (char) or index (int) of the first year column.
#' @param year_col2 The name (char) or index (int) of the second year column to be subtracted from.
#' @param year_col1_name The name alias (char) of the first year column year_col1.
#' @param year_col2_name The name alias (char) of the second year column year_col2.
#' @param year_min The minimum year (int) to calculate differences for.
#' @param year_max The maximum year (int) to calculate differences for.
#' @param title The title (char) of the plot.
#' @param title_n Set to TRUE to add the number of total segments considered.
#' @param x_title The title (char) of the x-axis.
#' @param y_title The title (char) of the y-axis.
#' @param x_breaks The number (int) of breaks to show on the x-axis. Set to FALSE to let ggplot automatically decide.
#' @param x_perc Set to TRUE to show proportions and FALSE to show counts.
#' @param out_data Set to TRUE to return a list
#' 
#' @return A ggplot of yearly differences (year_col2 - year_col1), displaying the proportion of rows for each difference in years. If `out_data` is TRUE then returns a list with keys `data` representing the data used for plotting and `plot` with the ggplot object.
#' @export
#'
plot_yearly_diff <- function(
		df,
		year_col1 = "install_year",
		year_col2 = "verify_install_year",
		year_col1_name = "City Year",
		year_col2_name = "Verified Year",
		year_min = settings$year_min,
		year_max = settings$year_max,
		title = sprintf(
			"Difference in Years, Comparing %s and %s",
			year_col1_name,
			year_col2_name
		),
		title_n = TRUE,
		x_title = sprintf(
			"Difference in Years (%s - %s)",
			year_col2_name,
			year_col1_name
		),
		y_title = "Proportion of Total Segments",
		x_breaks = 15,
		x_perc = TRUE,
		out_data = FALSE
) {
	
	# Filter for comparable rows only
	pdata <- df %>% filter(
		!is.na(.data[[year_col1]]) & !is.na(.data[[year_col2]])
	)
	
	# Filter within min year
	if (year_min) {
		pdata <- pdata %>% filter(
			.data[[year_col2]] > year_min
		)
	}
	
	# Filter within max year
	if (year_max) {
		pdata <- pdata %>% filter(
			.data[[year_col2]] <= year_max
		)
	}
	
	# Add n to title
	if (title_n) {
		title <- sprintf("%s (n=%s)", title, nrow(pdata))
	}
	
	# Calc yearly diff
	pdata <- pdata %>%
		mutate(year_diff = .data[[year_col2]] - .data[[year_col1]]) %>%
		count(year_diff) %>%
		mutate(n_perc = (n / sum(n)) * 100)
	
	# Set to proportions or counts
	pdata$y <- if (x_perc) pdata$n_perc else pdata$n
	
	# Plot yealy diffs
	out <- pdata %>% 
		ggplot(aes(
			x = year_diff,
			y = y
		)) +
		geom_bar(
			stat = "identity",
			color = "#332a94",
			fill = "#c3d5e4",
			width = 1
		) +
		labs(
			title = title,
			x = x_title,
			y = y_title
		) +
		theme(
			plot.title = element_text(size = 12)
		)
	
	# Add percentage sign if percentages
	if (x_perc) {
		out <- out +
			scale_y_continuous(
				label = scales::label_number(suffix = "%")
			)
	}
	
	# Set x interval breaks
	if (x_breaks) {
		out <- out + scale_x_continuous(
			breaks = scales::breaks_pretty(x_breaks)
		)
	}
	
	# Returns ggplot obj or list
	if (out_data) {
		out <- list(
			data = pdata,
			plot = out
		)
	}
	return(out)
}

```

## Function 5: filter_criteria

Fitler for segment inclusion criteria

This function applies segment inclusion critieria to a list of data.frames. Optionally creates a data.frame of counts, segment lengths, and other exclusions (duplicates, misclassifications) per inclusion criteria step along with a list of the data.frames after applying the inclusion criteria.

```{r}

#' Filter for Segment Inclusion Criteria
#'
#' This function applies segment inclusion critieria to a list of data.frames. Optionally creates a data.frame of counts, segment lengths, and other exclusions (duplicates, misclassifications) per inclusion criteria step along with a list of the data.frames after applying the inclusion criteria.
#'
#' @param criteria_data A list of lists, where each list contains the following structure defining the segment inclusion criteria for each city:
#' \itemize{
#'	\item \code{city}: the name (char) of the city (required).
#'	\item \code{data}: the data.frame containing road segments and applicable columns for inclusion criteria filtering (required).
#'	\item \code{data_date}: the date (char) that the data was acquired.
#'	\item \code{infra_col}: the column name (char) of the column containing the dedicated cycling infrastructure types to filter.
#'	\item \code{infra_filter}: the vector of characters of dedicated cycling infrastructure types to include.
#'	\item \code{road_col}: the column name (char) of the column containing the road location types to filter.
#'	\item \code{road_filter}: the vector of characters of road location types to exclude.
#'	\item \code{status_col}: the column name (char) of the column containing the inactive road status types to filter.
#'	\item \code{status_filter}: the vector of characters of inactive road status types to include.
#'	\item \code{geom_col}: the column name (char) of the column containing geometries.
#'  \item \code{geom_unit}: the unit measure (char) of the geometry 
#'	\item \code{geom_filter}: Set to TRUE to filter for null and duplicate geometries.
#'	\item \code{misclass_col}: the column name (char) of the column containing misclassification types to filter.
#'	\item \code{misclass_filter}: the vector of characters indicating non-misclassified rows of data to include. Usually set to c("NA", NA) to indicate that the row is not misclassified.
#'  \item \code{noverify_col}: the column containing infrastructure install types (char) that are not verified. This does not filter the data, but calculates and adjusts for the rows and road lengths of non-verified segments.
#'. \item \code{noverify_filter}: the vector of characters of non-verified infrastructure install types from the city. This does not filter the data, but calculates and adjusts for the rows and road lengths of non-verified segments.
#' }
#' @param len_func A function to apply to road length calculations. The default is a function that converts from meters to km.
#' 
#' @return A list of lists, where each list has keys and values from \code{criteria_data}, and the following additional keys:
#' \itemize{
#'	\item \code{data_filter}: the data.frame after filtering for segment inclusion criteria (required).
#'	\item \code{infra_filter_applied}: TRUE if dedicated cycling infrastructure filter was applied and FALSE otherwise (required).
#'	\item \code{infra_filter_n}: total rows (numeric) after filtering for dedicated cycling infrastructure using \code{infra_filter} (required).
#'	\item \code{infra_filter_len}: total road length (numeric) after filtering for dedicated cycling infrastructure using \code{infra_filter} (required).
#'	\item \code{infra_filter_nx}: total rows (numeric) affected by filtering for dedicated cycling infrastructure using \code{infra_filter} (required).
#'	\item \code{infra_filter_lenx}: total road length (numeric) affected by filtering for dedicated cycling infrastructure using \code{infra_filter} (required).
#'	\item \code{road_filter_applied}: TRUE if road location filter was applied and FALSE otherwise (required).
#'	\item \code{road_filter_n}: total rows (numeric) after filtering for road location using \code{infra_filter} (required).
#'	\item \code{road_filter_len}: total road length (numeric) after filtering for road location using \code{infra_filter} (required).
#'	\item \code{road_filter_nx}: total rows (numeric) affected by filtering for road location using \code{infra_filter} (required).
#'	\item \code{road_filter_lenx}: total road length (numeric) affected by filtering for road location using \code{infra_filter} (required).
#'	\item \code{status_filter_applied}: TRUE if inactive road status filter was applied and FALSE otherwise (required).
#'	\item \code{status_filter_n}: total rows (numeric) after filtering for inactive road status using \code{status_filter} (required).
#'	\item \code{status_filter_len}: total road length (numeric) after filtering for inactive road status using \code{status_filter} (required).
#'	\item \code{status_filter_nx}: total rows (numeric) affected by filtering for inactive road status using \code{status_filter} (required).
#'	\item \code{status_filter_lenx}: total road length (numeric) affected by filtering for inactive road status using \code{status_filter} (required).
#'	\item \code{geom_filter_null_applied}: TRUE if null geometries filter was applied and FALSE otherwise (required).
#'	\item \code{geom_filter_null_n}: total rows (numeric) after filtering for null geometries (required).
#'	\item \code{geom_filter_null_len}: total road length (numeric) after filtering for null geometries (required).
#'	\item \code{geom_filter_null_nx}: total rows (numeric) affected by filtering for null geometries (required).
#'	\item \code{geom_filter_null_lenx}: total road length (numeric) affected by filtering for null geometries (required).
#'	\item \code{geom_filter_dup_applied}: TRUE if duplicate geometries filter was applied and FALSE otherwise (required).
#'	\item \code{geom_filter_dup_n}: total rows (numeric) after filtering for duplicate geometries (required).
#'	\item \code{geom_filter_dup_len}: total road length (numeric) after filtering for duplicate geometries (required).
#'	\item \code{geom_filter_dup_nx}: total rows (numeric) affected by filtering for duplicate geometries (required).
#'	\item \code{geom_filter_dup_lenx}: total road length (numeric) affected by filtering for duplicate geometries (required).
#'	\item \code{elig_n}: total rows (numeric) after the above filters eligible for data entry and screening (required).
#'	\item \code{elig_len}: total road length (numeric) after the above filters eligible for data entry and screening (required).
#'	\item \code{misclass_filter_applied}: TRUE if null misclassifications filter was applied and FALSE otherwise (required).
#'	\item \code{misclass_filter_n}: total rows (numeric) after filtering misclassifications using \code{misclass_filter} (required).
#'	\item \code{misclass_filter_len}: total road length (numeric) after filtering misclassifications using \code{misclass_filter} (required).
#'	\item \code{misclass_filter_nx}: total rows (numeric) affected by filtering misclassifications using \code{misclass_filter} (required).
#'	\item \code{misclass_filter_lenx}: total road length (numeric) affected by misclassifications using \code{misclass_filter} (required).
#'	\item \code{misclass_filter_uniq_n}: a data.frame of total rows for each misclassification type.
#'	\item \code{misclass_filter_uniq_len}: a data.frame of total road lengths for each misclassification type.
#'	\item \code{noverify_filter_applied}: TRUE if non-verified infrastructure filter was calculated and FALSE otherwise (required).
#'	\item \code{noverify_filter_nx}: total rows (numeric) of non-verified infrastructure from \code{noverify_filter} (required).
#'	\item \code{noverify_filter_lenx}: total road length (numeric) affected by non-verified infrastructure using \code{noverify_filter} (required).
#'	\item \code{incl_n}: final total rows (numeric) after the above filters (required).
#'	\item \code{incl_len}: final total road length (numeric) after the above filters (required).
#' }
#' @export
#'
filter_criteria <- function(
	criteria_data,
	len_func = function (x) as.numeric(x) / 1000
) {
	
	# Apply criteria to list and track counts and lengths
	out <- criteria_data
	for (i in 1:length(criteria_data)) {
		
		# Get criteria data
		x <- criteria_data[[i]]
		df <- x$data
		city <- x$city
		
		# Set initial apply status for filters
		out[[city]]$infra_filter_applied <- FALSE
		out[[city]]$road_filter_applied <- FALSE
		out[[city]]$status_filter_applied <- FALSE
		out[[city]]$geom_filter_null_applied <- FALSE
		out[[city]]$geom_filter_dup_applied <- FALSE
		out[[city]]$misclass_filter_applied <- FALSE
		out[[city]]$noverify_filter_applied <- FALSE
		
		# Count/len initial
		out[[city]]$data_n <- nrow(df)
		out[[city]]$data_len <- len_func(sum(st_length(df[[x$geom_col]]), na.rm = TRUE))
		
		# Filter for dedicated cycling infra
		if (all(c("infra_col", "infra_filter") %in% names(x))) {
			
			# Apply ded cyc infra filter
			df <- df %>%
				filter(.data[[x$infra_col]] %in% x$infra_filter)
			
			# Set ded cyc infra filter status
			out[[city]]$infra_filter_applied <- TRUE
		}
		
		# Count/len ded cyc infra filter
		out[[city]]$infra_filter_n <- nrow(df)
		out[[city]]$infra_filter_len <- len_func(sum(st_length(df[[x$geom_col]]), na.rm = TRUE))
		
		# Count/len affected by ded cyc infra filter
		out[[city]]$infra_filter_nx <- out[[city]]$data_n - out[[city]]$infra_filter_n
		out[[city]]$infra_filter_lenx <- out[[city]]$data_len - out[[city]]$infra_filter_len
		
		# Filter for road location
		if (all(c("road_col", "road_filter") %in% names(x))) {
			
			# Apply road filter
			df <- df %>%
				filter(!.data[[x$road_col]] %in% x$road_filter)
			
			# Set road filter status
			out[[city]]$road_filter_applied <- TRUE
		}
		
		# Count/len road filter
		out[[city]]$road_filter_n <- nrow(df)
		out[[city]]$road_filter_len <- len_func(sum(st_length(df[[x$geom_col]]), na.rm = TRUE))
		
		# Count/len affected by road filter
		out[[city]]$road_filter_nx <- out[[city]]$infra_filter_n - out[[city]]$road_filter_n
		out[[city]]$road_filter_lenx <- out[[city]]$infra_filter_len - out[[city]]$road_filter_len
		
		# Filter for status
		if (all(c("status_col", "status_filter") %in% names(x))) {
			
			# Apply status filter
			df <- df %>%
				filter(!.data[[x$status_col]] %in% x$status_filter)
			
			# Set status filter status
			out[[city]]$status_filter_applied <- TRUE
		}
		
		# Count/len status filter
		out[[city]]$status_filter_n <- nrow(df)
		out[[city]]$status_filter_len <- len_func(sum(st_length(df[[x$geom_col]]), na.rm = TRUE))
		
		# Count/len affected by status filter
		out[[city]]$status_filter_nx <- out[[city]]$road_filter_n - out[[city]]$status_filter_n
		out[[city]]$status_filter_lenx <- out[[city]]$road_filter_len - out[[city]]$status_filter_len
		
		# Filter for null geoms
		if (all(c("geom_col", "geom_filter") %in% names(x))) {
			
			# Apply null geom filter
			df <- df %>%
				filter(!is.na(.data[[x$geom_col]]))
			
			# Set dup geom filter status
			out[[city]]$geom_filter_null_applied <- TRUE
		}
		
		# Count/len null geom filter
		out[[city]]$geom_filter_null_n <- nrow(df)
		out[[city]]$geom_filter_null_len <- len_func(sum(st_length(df[[x$geom_col]]), na.rm = TRUE))
		
		# Count/len affected by null geom filter
		out[[city]]$geom_filter_null_nx <- out[[city]]$status_filter_n - out[[city]]$geom_filter_null_n
		out[[city]]$geom_filter_null_lenx <- out[[city]]$status_filter_len - out[[city]]$geom_filter_null_len
			
		# Filter for dup geoms
		if (all(c("geom_col", "geom_filter") %in% names(x))) {
			
			# Apply dup geom filter
			df <- df %>%
				distinct(.data[[x$geom_col]], .keep_all = TRUE)
			
			# Set dup geom filter status
			out[[city]]$geom_filter_dup_applied <- TRUE
		}
		
		# Count/len dupl geom filter
		out[[city]]$geom_filter_dup_n <- nrow(df)
		out[[city]]$geom_filter_dup_len <- len_func(sum(st_length(df[[x$geom_col]]), na.rm = TRUE))
		
		# Count/len affected by dupl geom filter
		out[[city]]$geom_filter_dup_nx <- out[[city]]$geom_filter_null_n - out[[city]]$geom_filter_dup_n
		out[[city]]$geom_filter_dup_lenx <- out[[city]]$geom_filter_null_len - out[[city]]$geom_filter_dup_len
		
		# Calculate noverify segments
		if (all(c("noverify_col", "noverify_filter") %in% names(x))) {
			
			# Apply noverify filter separately
			df_noverify <- df %>%
				filter(!is.na(.data[[x$noverify_col]]))
			
			# Set noverify filter status
			out[[city]]$noverify_filter_applied <- TRUE
			
			# Count/len of noverify segments
			out[[city]]$noverify_filter_nx <- df_noverify %>% nrow
			out[[city]]$noverify_filter_lenx <- len_func(sum(st_length(df_noverify[[x$geom_col]]), na.rm = TRUE))
			
		} else {
			
			# Set to 0 if all segments are verified
			out[[city]]$noverify_filter_nx <- len_func(as_units(0, "meters"))
			out[[city]]$noverify_filter_lenx <- len_func(as_units(0, "meters"))
		}
		
		# Count/len eligible
		out[[city]]$elig_n <- nrow(df) - out[[city]]$noverify_filter_nx
		out[[city]]$elig_len <- len_func(sum(st_length(df[[x$geom_col]]), na.rm = TRUE))
		
		# Filter for misclass
		if (all(c("misclass_col", "misclass_filter") %in% names(x))) {
			
			# Count/len misclass groups
			out[[city]]$misclass_filter_uniq_n <- df %>%
				filter(!is.na(.data[[x$misclass_col]])) %>%
				count(.data[[x$misclass_col]]) %>%
				as_tibble
			out[[city]]$misclass_filter_uniq_len <- df %>%
				filter(!.data[[x$misclass_col]] %in% x$misclass_filter) %>%
				group_by(.data[[x$misclass_col]]) %>%
				summarize(len = len_func(sum(st_length(.data[[x$geom_col]]), na.rm = TRUE))) %>%
				as_tibble
			
			# Apply misclass filter
			df <- df %>%
				filter(.data[[x$misclass_col]] %in% x$misclass_filter)
			
			# Set misclass filter status
			out[[city]]$misclass_filter_applied <- TRUE
		}
		
		# Count/len misclass filter
		out[[city]]$misclass_filter_n <- nrow(df) - out[[city]]$noverify_filter_nx
		out[[city]]$misclass_filter_len <- len_func(sum(st_length(df[[x$geom_col]]), na.rm = TRUE)) - out[[city]]$noverify_filter_lenx
		
		# Count/len affected by misclass filter
		out[[city]]$misclass_filter_nx <- out[[city]]$elig_n - out[[city]]$misclass_filter_n
		out[[city]]$misclass_filter_lenx <- out[[city]]$elig_len - out[[city]]$misclass_filter_len
		
		# Count/len eligible
		out[[city]]$incl_n <- nrow(df) - out[[city]]$noverify_filter_nx
		out[[city]]$incl_len <- len_func(sum(st_length(df[[x$geom_col]]), na.rm = TRUE)) - out[[city]]$noverify_filter_lenx
		
		# Save filtered data
		out[[city]]$data_filter <- df
	}
	return(out)
}
```

## Function 5a: diag_criteria

Diagram the segment inclusion criteria results.

This function draws a flow diagram of overall methods for segment inclusion criteria using output from `filter_criteria`.

```{r}

#' Diagram the segment inclusion criteria results
#' 
#' This function draws a flow diagram of overall methods for segment inclusion criteria using output from \code{\link{filter_criteria}}.
#'
#' @param criteria_data A list of lists in the structure of the output from \code{\link{filter_criteria}}.
#' @param note A note (char) to display at the end of the diagram.
#' @return A \code{\link[DiagrammeR]{grViz}} object.
#' @export
#'
diag_criteria <- function(
		criteria_data,
		note = "*Denotes segments misclassified as eligible after screening<br/>**Local Street Bikeways (LSB) were included but not screened"
) {
	
	# Diag settings
	diag_settings <- "
		rankdir = LR
		node[
			shape = box,
			width = 2.75,
			height = 1.65,
			style = filled,
			fillcolor = white,
			penwidth = 1.5,
			fontname = 'Arial'
		]
		edge[
			arrowhead = vee,
			arrowtail = vee
		]
		layout = neato
	"
	
	# Top header row
	row_top <- "
		open_data[
			label = 'Open Data',
			height = 0.5,
			fillcolor = '#d7e9fe',
			pos = '0,1!'
		]
		elig_data[
			label = 'Eligible Segments',
			height = 0.5,
			fillcolor = '#d7e9fe',
			pos = '3.25,1!'
		]
		incl_data[
			label = 'Inclusions',
			height = 0.5,
			fillcolor = '#d7e9fe',
			pos = '6.5,1!'
		]
	"
	
	# Create template for row addition
	row_temp <- "
		open{i}[
			label = <<b>{city}</b><br/>N = {open_n} Segments<br/><i>({open_len})<br/>Downloaded: {open_date}</i>>,
			pos = '0,{y}!'
		]
		
		elig{i}[
			label = <n = {elig_n} Segments<br/>({elig_len})<i><br/><b>Exclusions</b>{elig_inelig}{elig_dup}{elig_poly}</i>>,
			pos = '3.25,{y}!'
		]
		
		incl{i}[
			label = <n = {incl_n} Segments<br/><i>({incl_len}){noverify}<br/><b>Exclusions</b>{incl_miss}{incl_dup}</i>>,
			pos = '6.5,{y}!'
		]
		
		open{i} -> elig{i} -> incl{i}
	"
	
	# Generate row additions per city
	y <- -0.21
	y_gap <- 1.85
	row_adds <- ""
	for (i in 1:length(criteria_data)) {
		
		# Vars per city
		criteria <- criteria_data[[i]]
		
		# Generate geom filter dup info
		elig_dup <- ""
		if (criteria$geom_filter_dup_nx > 0) {
			elig_dup <- glue(
				"<br/>Duplicates: n = {n}",
				n = criteria$geom_filter_dup_nx
			)
		}
		
		# Generate geom filter null info
		elig_poly <- ""
		if (criteria$geom_filter_null_nx > 0) {
			elig_poly <- glue(
				"<br/>No Polyline Data: n = {n}",
				n = criteria$geom_filter_null_nx
			)
		}
		
		# Generate inelig info
		elig_inelig <- glue(
			"<br/>Ineligible: n = {n}",
			n = criteria[["infra_filter_nx"]] + criteria[["status_filter_nx"]] + criteria[["road_filter_nx"]]
		)
		
		# Generate noverify info
		noverify <- ""
		if (criteria$noverify_filter_applied) {
			noverify <- glue(
				"<br/>**Screened: n = {n}<br/>**Not screened: n = {nx}",
				n = criteria$elig_n,
				nx = criteria$noverify_filter_nx
			)
		}
		
		# Generate incl info
		incl_miss <- glue(
			"<br/>*Misclassifications: n = {n}",
			n = criteria[["misclass_filter_nx"]]
		)
		
		# Road length unit
		if ("geom_unit" %in% names(criteria)) {
			len_unit <- criteria$geom_unit
		} else {
			len_unit = "meters"
		}
		
		# Generate single row addition
		row_adds <- paste0(row_adds, glue(
			row_temp,
			i = i,
			y = y,
			city = str_to_title(criteria[["city"]]),
			open_n = criteria[["data_n"]],
			open_len = paste(round(criteria[["data_len"]], 1), len_unit),
			open_date = criteria[["data_date"]],
			elig_n = criteria$elig_n + criteria$noverify_filter_nx,
			elig_len = paste(round(criteria[["elig_len"]], 1), len_unit),
			elig_inelig = elig_inelig,
			elig_dup = elig_dup,
			elig_poly = elig_poly,
			incl_n = criteria[["incl_n"]] + criteria$noverify_filter_nx,
			incl_len = paste(round(criteria[["incl_len"]] + criteria$noverify_filter_lenx, 1), len_unit),
			incl_miss = incl_miss,
			incl_dup = "",
			noverify = noverify
		))
		
		# Move row below
		y <- y - y_gap
	}
	
	# Filter and screening lines
	line_filter <- glue("
		filter1[
			label = 'Filtering',
			height = 0.25,
			shape = plaintext,
			style='', pos = '1.6,1.425!'
		]
		filter2[
			style = invis,
			pos = '1.6,{y}!'
		]
		filter1 -> filter2 [style = dashed, dir = none, color = '#b0b0b0']
	", y = y - -0.96)
	line_screen <- glue("
		screen1[
			label = 'Screening',
			height = 0.25,
			shape = plaintext,
			style='', pos = '4.85,1.425!'
		]
		screen2[
			style = invis,
			pos = '4.85,{y}!'
		]
		screen1 -> screen2 [style = dashed, dir = none, color = '#b0b0b0']
	", y = y - -0.96)
	
	# Bottom note
	note_bottom <- glue("
		note[
			label=<<i>{text}</i>>,
			style = '',
			shape = plaintext,
			fontsize = 12,
			pos = '3.25,{y}!'
		]
	", text = note, y = y - -0.69)
	
	# Generate graphviz diag
	out <- grViz(paste0(
		"digraph {",
		diag_settings,
		row_top,
		row_adds,
		line_filter,
		line_screen,
		note_bottom,
		"}"
	))
	return(out)
}
```

## Function 5b: diag_criteria_details

Diagram the segment inclusion criteria results in detail.

This function draws a flow diagram of detailed methods for segment inclusion criteria using output from `filter_criteria`.

```{r}
#' Diagram the segment inclusion criteria results in detail
#' 
#' This function draws a flow diagram of detailed methods for segment inclusion criteria using output from \code{\link{filter_criteria}}.
#'
#' @param criteria_data A list of lists in the structure of the output from \code{\link{filter_criteria}}.
#' @param city The city (char) to create the diagram for. If `NULL`, this function produces a list of diagrams where keys are the city name and values are the diagrams.
#' @param out_render Set to TRUE to render the diagram and return \code{\link[DiagrammeR]{grViz}} objects or FALSE to return the text used to generate the diagram.
#' @return A list of \code{\link[DiagrammeR]{grViz}} objects if `city` is `NULL`, or a single \code{\link[DiagrammeR]{grViz}} if `city` is provided. The \code{\link[DiagrammeR]{grViz}} objects become text (char) if `out_render` is `FALSE`.
#' @export
#'
diag_criteria_details <- function(criteria_data, city = NULL, out_render = TRUE) {
	
	# Filter for city if avail
	if (!is.null(city)) {
		criteria_data <- criteria_data[sapply(criteria_data, function (x) x$city == city)]
	}
	
	# Generate diagrams for each city
	out <- list()
	for (i in 1:length(criteria_data)) {
		
		# Diag vars
		criteria <- criteria_data[[i]]
		x_edge <- -4
		
		# Diag settings
		diag_settings <- "
			rankdir = TB
			node[
				shape = box
				width = 10
				height = 1.8
				style = filled
				fillcolor = white
				penwidth = 1.5
				fontsize = 16
				fontname = 'Arial'
				margin = 0.25
			]
			edge[
				arrowhead = vee,
				arrowtail = vee
			]
			layout = neato
		"
		
		# Step 1 identification
		s1 <- glue("
			id_title[
				label = <<b>Identification</b>>
				pos = '-8.5,0!'
				width = 2
				height = 1.9
				fillcolor = '#d7e9fe'
				style = 'rounded,filled'
			]
			id[
				label = 'Shapefile from: {url}\\lDownloaded: {date}\\lN = {n} Segments\\l'
				pos = '0,0!'
				width = 14
			]
			
			id_top[
				style = invis
				pos = '{x},0!'
			]
			id_bot[
				style = invis
				pos = '{x},-2.25!'
			]
			id_top -> id_bot
		",
			url = criteria$data_url,
			date = criteria$data_date,
			n = criteria$data_n,
			x = x_edge
		)
		
		# Step 2 vars
		fi <- 0
		y <- -0
		s2 <- ""
		
		# Step 2 filtering infra
		if (criteria$infra_filter_applied) {
			fi <- fi + 1
			y <- y - 2.25
			s2 <- glue("
				{s2}

				filter{fi}[
					label = 'Filter for Dedicated Cycling Infrastructure\\l{column} in {filter}\\l(n = {n})\\l'
					pos = '-2,{y}!'
				]
				filter{fi}x[
					label = 'Segments Excluded\\l(n = {nx})\\l'
					pos = '5.5,{y}!'
					width = 3
				]
				filter{fi} -> filter{fi}x
				
				filter{fi}_top[
					style = invis
					pos = '{x},{y}!'
				]
				filter{fi}_bot[
					style = invis
					pos = '{x},{y - 2.25}!'
				]
				filter{fi}_top -> filter{fi}_bot
			",
				column = criteria$infra_col,
				filter = str_replace_all(
					str_wrap(
						paste0(
							criteria$infra_filter,
							collapse = ", "
						),
						width = 83
					),
					"[\r\n]",
					"\\\\l"
				),
				n = criteria$infra_filter_n,
				nx = criteria$infra_filter_nx,
				fi = fi,
				y = y,
				x = x_edge,
				s2 = s2
			)
		}
		
		# Step 2 filtering road
		if (criteria$road_filter_applied) {
			fi <- fi + 1
			y <- y - 2.25
			s2 <- glue("
				{s2}

				filter{fi}[
					label = 'Filter for Infrastructure Located on Roadway\\l{column} != {filter}\\l(n = {n})\\l'
					pos = '-2,{y}!'
				]
				filter{fi}x[
					label = 'Segments Excluded\\l(n = {nx})\\l'
					pos = '5.5,{y}!'
					width = 3
				]
				filter{fi} -> filter{fi}x
				
				filter{fi}_top[
					style = invis
					pos = '{x},{y}!'
				]
				filter{fi}_bot[
					style = invis
					pos = '{x},{y - 2.25}!'
				]
				filter{fi}_top -> filter{fi}_bot
			",
				column = criteria$road_col,
				filter = str_replace_all(
					str_wrap(
						paste0(
							criteria$road_filter,
							collapse = ", "
						),
						width = 83
					),
					"[\r\n]",
					"\\\\l"
				),
				n = criteria$road_filter_n,
				nx = criteria$road_filter_nx,
				fi = fi,
				y = y,
				x = x_edge,
				s2 = s2
			)
		}

		# Step 2 filtering status
		if (criteria$status_filter_applied) {
			fi <- fi + 1
			y <- y - 2.25
			s2 <- glue("
				{s2}

				filter{fi}[
					label = 'Filter for Active Infrastructure Status\\l{column} != {filter}\\l(n = {n})\\l'
					pos = '-2,{y}!'
				]
				filter{fi}x[
					label = 'Segments Excluded\\l(n = {nx})\\l'
					pos = '5.5,{y}!'
					width = 3
				]
				filter{fi} -> filter{fi}x
				
				filter{fi}_top[
					style = invis
					pos = '{x},{y}!'
				]
				filter{fi}_bot[
					style = invis
					pos = '{x},{y - 2.25}!'
				]
				filter{fi}_top -> filter{fi}_bot
			",
				column = criteria$status_col,
				filter = str_replace_all(
					str_wrap(
						paste0(
							criteria$status_filter,
							collapse = ", "
						),
						width = 83
					),
					"[\r\n]",
					"\\\\l"
				),
				n = criteria$status_filter_n,
				nx = criteria$status_filter_nx,
				fi = fi,
				y = y,
				x = x_edge,
				s2 = s2
			)
		}

		# Step 2 filtering null geom
		if (criteria$geom_filter_null_applied) {
			fi <- fi + 1
			y <- y - 2.25
			s2 <- glue("
				{s2}

				filter{fi}[
					label = 'Filter for Null Geometry\\l{column} is not null\\l(n = {n})\\l'
					pos = '-2,{y}!'
				]
				filter{fi}x[
					label = 'Segments Excluded\\l(n = {nx})\\l'
					pos = '5.5,{y}!'
					width = 3
				]
				filter{fi} -> filter{fi}x
				
				filter{fi}_top[
					style = invis
					pos = '{x},{y}!'
				]
				filter{fi}_bot[
					style = invis
					pos = '{x},{y - 2.25}!'
				]
				filter{fi}_top -> filter{fi}_bot
			",
				column = criteria$geom_col,
				n = criteria$geom_filter_null_n,
				nx = criteria$geom_filter_null_nx,
				fi = fi,
				y = y,
				x = x_edge,
				s2 = s2
			)
		}
		
		# Step 2 filtering dup geom
		if (criteria$geom_filter_dup_applied) {
			fi <- fi + 1
			y <- y - 2.25
			s2 <- glue("
				{s2}

				filter{fi}[
					label = 'Filter for Duplicate Geometry\\l{column} is not duplicated\\l(n = {n})\\l'
					pos = '-2,{y}!'
				]
				filter{fi}x[
					label = 'Segments Excluded\\l(n = {nx})\\l'
					pos = '5.5,{y}!'
					width = 3
				]
				filter{fi} -> filter{fi}x
				
				filter{fi}_top[
					style = invis
					pos = '{x},{y}!'
				]
				filter{fi}_bot[
					style = invis
					pos = '{x},{y - 2.25}!'
				]
				filter{fi}_top -> filter{fi}_bot
			",
				column = criteria$geom_col,
				n = criteria$geom_filter_dup_n,
				nx = criteria$geom_filter_dup_nx,
				fi = fi,
				y = y,
				x = x_edge,
				s2 = s2
			)
		}
		
		# Step 2 filtering
		s2 <- glue("
			filter_title[
				label = <<b>Filtering</b>>
				pos = '-8.5,{y}!'
				width = 2
				height = {h}
				fillcolor = '#d7e9fe'
				style = 'rounded,filled'
			]
			{s2}
		",
			h = (fi * 2.1),
			fi = fi,
			y = y + if (fi == 1) 0 else (((fi -1) / 2) * 2.25),
			s2 = s2
		)
		
		# Step 3 eligible
		y <- y - 2.25
		s3 <- glue("
			elig_title[
				label = <<b>Eligible</b>>
				pos = '-8.5,{y}!'
				width = 2
				height = 1.9
				fillcolor = '#d7e9fe'
				style = 'rounded,filled'
			]
			elig[
				label = 'Segments Included for Data Entry and Screening\\l(n = {n})\\l'
				pos = '0,{y}!'
				width = 14
			]
			
			elig_top[
				style = invis
				pos = '{x},{y}!'
			]
			elig_bot[
				style = invis
				pos = '{x},{y - 2.25}!'
			]
			elig_top -> elig_bot
		",
			n = criteria$elig_n + criteria$noverify_filter_nx,
			y = y,
			x = x_edge
		)
		
		# Step 4 Screening
		s4 <- ""
			
		# Step 4 title
		y <- y - 2.65
		s4 <- glue("
			screen_title[
				label = <<b>Screening</b>>
				pos = '-8.5,{y}!'
				width = 2
				height = 2.55
				fillcolor = '#d7e9fe'
				style = 'rounded,filled'
			]
		",
			n = criteria$misclass_n,
			y = y
		)
		
		# Step 4 noverify
		misclass_noverify <- ""
		if (criteria$noverify_filter_nx > 0) {
			misclass_noverify <- glue(
				"{n} screened, {nx} not screened\\l",
				n = criteria$misclass_filter_n,
				nx = criteria$noverify_filter_nx
			)
		}
		
		# Step 4 misclass
		s4 <- glue("
			{s4}
			
			screen[
				label = 'Exclusion of Misclassifications and\\lDuplicates following Screening\\l{column} != {filter}\\l{noverify}(n = {n})\\l'
				pos = '-4.5,{y}!'
				width = 5
				height = 2.5
			]
			screenx[
				label = '{misclass}'
				pos = '3,{y}!'
				width = 7.95
				height = 2.5
			]
			screen -> screenx
			
			screen_top[
				style = invis
				pos = '{x},{y - 0.35}!'
			]
			screen_bot[
				style = invis
				pos = '{x},{y - 2.75}!'
			]
			screen_top -> screen_bot
		",
			column = criteria$misclass_col,
			filter = str_replace_all(
				str_wrap(
					paste0(
						criteria$misclass_filter,
						collapse = ", "
					),
					width = 83
				),
				"[\r\n]",
				"\\\\l"
			),
			n = criteria$misclass_filter_n + criteria$noverify_filter_nx,
			noverify = misclass_noverify,
			misclass = paste0(
				"Misclassifications: ",
				criteria$misclass_filter_uniq_n[[1]],
				" (n = ",
				criteria$misclass_filter_uniq_n[[2]],
				")\\l",
				collapse = ""
			),
			y = y,
			x = x_edge,
			s4 = s4
		)
		
		# Step 5 noverify
		incl_noverify <- ""
		if (criteria$noverify_filter_nx > 0) {
			incl_noverify <- glue(
				"{n} verified, {nx} not verified\\l",
				n = criteria$incl_n,
				nx = criteria$noverify_filter_nx
			)
		}
		
		# Step 5 Inclusions
		y <- y - 2.75
		s5 <- glue("
			incl_title[
				label = <<b>Inclusions</b>>
				pos = '-8.5,{y}!'
				width = 2
				height = 1.9
				fillcolor = '#c8e29d'
				style = 'rounded,filled'
			]
			incl[
				label = '{verified}Inclusions\\l{noverify}(n = {n})\\l'
				pos = '0,{y}!'
				width = 14
			]
		",
			verified = if (criteria$noverify_filter_nx > 0) "Verified and Non-verified " else "Verified ",
			noverify = incl_noverify,
			n = criteria$incl_n + criteria$noverify_filter_nx,
			y = y
		)
		
		# Combine steps
		out[[criteria$city]] <- paste0(
			"digraph {\n",
			diag_settings,
			"\n",
			s1,
			"\n",
			s2,
			"\n",
			s3,
			"\n",
			s4,
			"\n",
			s5,
			"\n",
			"}"
		)
	}
	
	# Return diagrams or single diagram if city is given
	out <- if (length(out) > 1) out else out[[1]]
	out <- if (out_render) grViz(out) else out
	return(out)
}
```

## Function 6: prep_infra

Prepare Infrastructure Changes Data for Mapping.

```{r}
#' Prepare Infrastructure Changes Data for Mapping
#'
#' This function prepares city data in a list format for mapping infrastructure changes since a target year.
#'
#' @param map_list A list of lists, where each list contains the following structure defining the city mapping data and settings:
#' \itemize{
#'  \item \code{title}: the title (char) of the main city map.
#'	\item \code{data}: the sf data.frame containing road segments of the install, upgrade1, and upgrade2 years and types (required).
#'	\item \code{downtown_bbox}: a vector (numeric) containing the coordinates of the downtown region's bounding box in xmin, ymin, xmax, and ymax respectively.
#' }
#' @param year_since The year (numeric) since to examine infrastructure changes.
#' 
#' @return A list of lists, where each list has keys and values from \code{map_list}, and the following additional keys:
#' \itemize{
#'	\item \code{data_map}: a sf data.frame with an additional `changes` column indicating the infrastructure changes since the target `year_since`.
#'	\item \code{data_bbox}: a sf data.frame of the bounding box of `data_map`.
#'	\item \code{data_downtown}: Same as `data_map` except for the downtown region indicated by `downtown_bbox`.
#'	\item \code{data_downtown_bbox}: a sf data.frame of the bounding box of `data_downtown`.
#'	\item \code{map_colors}: the colors (char) for each of the infrastructure change categories.
#'	\item \code{map_column}: the column name (char) to be mapped
#'	\item \code{downtown_title}: the name (char) of the downtown subset map
#' }
#' @export
#'
prep_infra <- function(
		map_list,
		year_since = settings$infra_changes_year
) {
	
	# Create color palette
	colors <- c("green", "orange", "gray")
	names(colors) <- c(
		glue("New Infrastructure Since Jan. {year}", year = year_since), # green
		glue("Upgraded Infrastructure Since Jan. {year}", year = year_since), # orange
		"Unchanged Infrastructure" # gray
	)
	
	# Generate maps per city
	out <- map_list
	for (i in 1:length(map_list)) {
		
		# Get city vars
		city <- map_list[[i]]
		
		# Create downtown title if not given
		if (!"downtown_title" %in% names(city)) {
			id <- names(map_list)[[i]]
			downtown_title <- glue(
				"Downtown {id}",
				id = str_to_title(id)
			)
		} else {
			downtown_title <- city$downtown_title
		}
		
		# Create col to identify infra changes
		map_data <- city$data %>%
			mutate(
				changes = case_when(
			        (
			        	!is.na(verify_upgrade1_type) &
			        	!is.na(verify_upgrade1_year) &
			         	verify_upgrade1_year >= year_since
			        ) | (
			        	!is.na(verify_upgrade2_type) &
			        	!is.na(verify_upgrade2_year) &
			        	verify_upgrade2_year >= year_since
			        ) ~ glue(
			        	"Upgraded Infrastructure Since Jan. {year}",
			        	year = year_since
			        ),
			        !is.na(verify_install_type) &
			        !is.na(verify_install_year) &
			        verify_install_year >= year_since ~
			        glue(
			        	"New Infrastructure Since Jan. {year}",
			        	year = year_since
			        ),
			        .default = "Unchanged Infrastructure"
			    )
			)
		
		# Create bounding box for city
		city_bbox <- st_as_sfc(
			st_bbox(city$data, crs = 4326)
		)
		
		# Create bounding box for downtown region
		downtown_bbox <- st_as_sfc(
			st_bbox(city$downtown_bbox, crs = 4326)
		)
		
		# Subset data for downtown region
		submap_data <- map_data %>% st_crop(downtown_bbox)
		
		# Add prep data to cities list
		out[[i]]$data_map <- map_data
		out[[i]]$data_bbox <- city_bbox
		out[[i]]$data_downtown <- submap_data
		out[[i]]$data_downtown_bbox <- downtown_bbox
		out[[i]]$map_colors <- colors
		out[[i]]$map_column <- "changes"
		out[[i]]$downtown_title <- downtown_title
	}
	return(out)
}
```

## Function 6a: map_infra

Maps Infrastructure Changes.

Creates maps of infrastructure changes since a certain year for each city and their downtown region using output from `prep_map`.

```{r}

#' Map Infrastructure Changes
#'
#' This function maps infrastructure changes since a target year.
#'
#' @inheritParams prep_infra
#' 
#' @return A `patchwork` object of `ggplot` objects combined together to form the multiple maps in arranged on a layout.
#' @export
#'
map_infra <- function(
		map_list,
		year_since = settings$infra_changes_year,
		scale_prop = 0.35
) {
	
	# Prepare data for maps
	cities_prep <- prep_infra(map_list)
		
	# Generate maps per city
	out <- list()
	for (i in 1:length(cities_prep)) {
		
		# Get city vars
		city <- cities_prep[[i]]
		id <- names(cities_prep)[[i]]
		
		# Create base map for city and downtown map
		base_map <- ggplot() +
		    annotation_map_tile(
		    	zoomin = 1,
		    	type = "cartolight",
		    	cachedir = "../data/cache"
		    ) +
			annotation_north_arrow(
				width = unit(0.2, "cm"),
				height = unit(0.5, "cm"),
				location = "br"
			) +
			annotation_scale(
				location = "bl",
				style = "ticks",
				width_hint = scale_prop
			) +
			scale_color_manual(values = city$map_colors) +
			fixed_plot_aspect(ratio = 1.5) +
			theme_void()
		
		# Generate city map
		out[[id]] <- base_map +
			ggtitle(city$title) +
			layer_spatial(city$data_map, aes(color = .data[[city$map_column]])) +
			layer_spatial(city$data_downtown_bbox, color = "red", fill = NA, linewidth = 0.5) +
			guides(colour = guide_legend(
				override.aes = list(linewidth = 3)
			))
		
		# Generate downtown map
		out[[paste0(id, "_downtown")]] <- base_map +
			ggtitle(city$downtown_title) +
			layer_spatial(city$data_downtown, aes(color = .data[[city$map_column]])) +
			guides(color = "none")
	}
	
	# Combine maps into single layout
	out <- wrap_plots(out, ncol = 2) + 
		plot_layout(guides = "collect") &
		theme(
			legend.position = "bottom",
			legend.title = element_blank(),
			plot.title = element_text(
				size = 8,
				margin = margin(t = 8, b = -20, l = 8)
			),
			plot.margin = margin(t = 8, l = 0, r = 0),
			panel.border = element_rect(
				colour = "gray20",
				fill = NA,
				linewidth = 0.5
			)
		)
	return(out)
}
```

## Function 6b: map_infra_detail

Maps Infrastructure Changes in Detail.

Creates enlarged maps of infrastructure changes since a certain year for each city and their downtown region.

```{r}

#' Map Infrastructure Changes in Detail
#'
#' This function creates enlarged maps of infrastructure changes since a target year.
#'
#' @inheritParams prep_infra
#' @param city_key They city key (char) to map from `map_list`. If `NULL`, maps all cities and returns a list, otherwise if given, returns an item from the list (required).
#' @param map_inset Set to `TRUE` to create an inset map of the downtown region or `FALSE` to omit the inset map.
#' @param map_inset_position A named vector (numeric) containing four values indicating the position of the inset map with the names being `left`, `bottom`, `right`, and `top` aligned to the `full` area. See \link[patchwork]{inset_element}.
#' 
#' @param map_ratio The aspect ratio (numeric) of the map.
#' @param map_inset_ratio The aspect ratio (numeric) of the subset map.
#' @return A list of `patchwork` object of `ggplot` objects combined together to form the enlarged maps, where the keys are the names of the cities as in `map_list`. If `city_key` is provided, returns only one of the items from this list.
#' @export
#'
map_infra_detail <- function(
		map_list,
		city_key = NULL,
		map_inset = TRUE,
		map_inset_position = c(
			left = 0.6,
			bottom = 0.6,
			right = 1,
			top = 1
		),
		map_ratio = 1.75,
		map_inset_ratio = 2,
		year_since = settings$infra_changes_year,
		...
) {
	
	# Only map one city if given
	if (!is.null(city_key)) {
		map_list <- list(map_list[[city_key]])
		names(map_list) <- city_key
	}
	
	# Prepare data for maps
	cities_prep <- prep_infra(map_list)
		
	# Generate enlarged maps per city
	out <- list()
	for (i in 1:length(cities_prep)) {
		
		# Get city vars
		city <- cities_prep[[i]]
		id <- names(cities_prep)[[i]]
		
		# Create base map for city and downtown map
		base_map <- ggplot() +
		    annotation_map_tile(
		    	zoomin = 1,
		    	type = "cartolight",
		    	cachedir = "../data/cache"
		    ) +
			scale_color_manual(values = city$map_colors) +
			theme_void()
		
		# Generate city map
		if ("map_ratio" %in% city) {
			map_ratio <- city$map_ratio
		}
		city_map <- base_map +
			fixed_plot_aspect(ratio = map_ratio) +
			annotation_north_arrow(
				width = unit(0.2, "cm"),
				height = unit(0.5, "cm"),
				location = "br"
			) +
			annotation_scale(
				location = "bl",
				style = "ticks"
			) +
			layer_spatial(city$data_map, aes(color = .data[[city$map_column]])) +
			guides(colour = guide_legend(
				override.aes = list(linewidth = 3)
			)) +
		    theme(
		        legend.position = "bottom",
		        legend.title = element_blank(),
		        panel.border = element_rect(
		            colour = "gray20",
		            fill = NA,
		            linewidth = 0.5
		        )
		    )
		
		# Add inset map as downtown region
		map_inset <- if ("map_inset" %in% names(city)) city$map_inset else map_inset
		if (map_inset) {
			
			# Generate downtown map
			if ("map_inset_ratio" %in% city) {
				map_inset_ratio <- city$map_inset_ratio
			}
			downtown_map <- base_map +
				fixed_plot_aspect(ratio = map_inset_ratio) +
			    layer_spatial(city$data_downtown, aes(color = .data[[city$map_column]])) +
			    guides(color = "none") +
			    annotation_scale(
			        location = "tl",
			        style = "ticks"
			    ) +
			    theme(
			        panel.border = element_rect(
			            colour = "black",
			            fill = NA,
			            linewidth = 0.75
			        )
			    )
			
			# Create final map with inset
			if ("map_inset_position" %in% names(city)) {
				map_inset_position <- city$map_inset_position
			}
			out[[id]] <- city_map + inset_element(
				downtown_map,
				left = map_inset_position[["left"]],
				bottom = map_inset_position[["bottom"]],
				right = map_inset_position[["right"]],
				top = map_inset_position[["top"]],
				align_to = "full"
			)
			
		} else {
			
			# No inset for final map
			out[[id]] <- city_map
		}
	}
	
	# Return list of all city maps or single map if city_key given
	if (!is.null(city_key)) {
		out <- out[[city_key]]
	}
	return(out)
}
```

## Function 7: plot_yearly_change

Plot yearly adjusted road length changes by infrastructure type.

This function plots line charts of yearly road length changes by infrastructure types for a list of data.

```{r}

#' Plot Yearly Road Length Changes By Infrastructure Type
#' 
#' Creates line plots of road length changes by infrastructure type.
#'
#' @param df_list A list of lists, where each key is the title and each value contains a list with the following structure:
#' \itemize{
#'   \item \code{city}: the name (char) of the city
#'   \item \code{data}: data.frame containing the install and change years, type, and road segment lengths.
#'   \item \code{roadway_total}: the total roadway length if `rodway_per` is given. This is used as the denominator to normalize road lengths.
#'.  \item \code{roadway_per}: Number of units of total roadway length (numeric) to normalize by (e.g. 1000 means per 1000 km of roadway). Set to `NULL` or omit to disable normalization of road lengths.
#' }
#' @param len_title The title (char) for the road lengths.
#'
#' @return Multiple line ggplots of the cumulative yearly road length changes by infrastructure type combined with patchwork.
#' @export
#'
plot_yearly_change <- function(
		df_list,
		len_title = "Change in Infrastructure (per 1000 centreline-km of roadway)"
	) {
	
	# Process plot data for adj len including total
	pdata <- list()
	for (i in 1:length(df_list)) {
		
		# Get data and vars
		df <- df_list[[i]]$data
		city <- df_list[[i]]$city
		
		# Get roadway vars if exists
		roadway_per <- NULL
		roadway_total <- NULL
		if ("roadway_per" %in% names(df_list[[i]])) {
			roadway_per <- df_list[[i]]$roadway_per
		}
		if ("roadway_total" %in% names(df_list[[i]])) {
			roadway_total <- df_list[[i]]$roadway_total
		}
		
		# Filter for study year period
		df <- df %>% filter(
			verify_install_year >= settings$year_min &
			verify_install_year <= settings$year_max
		)
		
		# Filter for types
		df <- df %>% filter(
			verify_install_type %in% c("PL", "BUF", "PBL", "LSB") |
			verify_upgrade1_type %in% c("PL", "BUF", "PBL", "LSB") |
			verify_upgrade2_type %in% c("PL", "BUF", "PBL", "LSB")
		)
		
		# Calc infra per year
		pdata[[i]] <- calc_yearly_adj_len(df) %>%
			mutate(
				city = city
			)
		
		# Calc total without lsb infra per year
		pdata_nolsb <- pdata[[i]] %>% filter(
			type != "LSB"
		) %>% group_by(
			year
		) %>% summarize(
			adj_len = sum(adj_len, na.rm = TRUE)
		) %>% mutate(
			type = "TOTAL"
		) %>% mutate(
			city = city
		)
		
		# Calc total with lsb infra per year
		pdata_lsb <- pdata[[i]] %>% group_by(
			year
		) %>% summarize(
			adj_len = sum(adj_len, na.rm = TRUE)
		) %>% mutate(
			type = "TOTAL_LSB"
		) %>% mutate(
			city = city
		)
		
		# Add totals as rows
		pdata[[i]] <- pdata[[i]] %>% add_row(pdata_nolsb)
		pdata[[i]] <- pdata[[i]] %>% add_row(pdata_lsb)
		
		# Norm len if needed
		if (!is.null(roadway_per)) {
			pdata[[i]] <- pdata[[i]] %>% mutate(
				adj_len_norm = 
					(adj_len / roadway_total) * roadway_per
			)
		}
	}
	
	# Combine plot data for each city
	pdata <- bind_rows(pdata)
	
	# Create infra line plots
	p <- list()
	
		# Create total with lsb infra line plot
	p[[1]] <- pdata %>% filter(
		type == "TOTAL_LSB"
	) %>% group_by(year, city) %>% summarize(
		adj_len_norm = sum(adj_len_norm, na.rm = TRUE)
	) %>% group_by(city) %>% arrange(year) %>% mutate(
		change = adj_len_norm - lag(adj_len_norm)
	) %>% ggplot(aes(
		x = year,
		y = change,
		color = factor(city, levels = c(
			"Vancouver", "Calgary", "Toronto"
		))
	)) + geom_line(
		size = 0.75
	) + geom_point() + geom_vline(
		xintercept = 2019,
		linetype = "dashed",
		color = "gray25",
		size = 0.5
	) + ggtitle(
		bquote(underline(.("Total On-Street Cycling Infrastructure")))
	) + scale_y_continuous(
		label = scales::label_number(suffix = " km")
	) + scale_fill_discrete(
		breaks = rev(c("Vancouver", "Calgary", "Toronto"))
	) + scale_colour_manual(
		values = c("#546ca9", "#c5a43d", "#719d71")
	) + scale_x_continuous(
		breaks = seq(
			settings$year_min + 1,
			settings$year_max,
			by = 1
		),
		limits = c(
			settings$year_min + 1,
			settings$year_max
		)
	) + theme(
		axis.title.y = element_blank(),
		axis.title.x = element_blank(),
		legend.title = element_blank()
	)
	
	# Create total without lsb infra line plot
	p[[2]] <- pdata %>% filter(
		type == "TOTAL"
	) %>% group_by(year, city) %>% summarize(
		adj_len_norm = sum(adj_len_norm, na.rm = TRUE)
	) %>% group_by(city) %>% arrange(year) %>% mutate(
		change = adj_len_norm - lag(adj_len_norm)
	) %>% ggplot(aes(
		x = year,
		y = change,
		color = factor(city, levels = c(
			"Vancouver", "Calgary", "Toronto"
		))
	)) + geom_line(
		size = 0.75
	) + geom_point() + geom_vline(
		xintercept = 2019,
		linetype = "dashed",
		color = "gray25",
		size = 0.5
	) + ggtitle(
		bquote(underline(.("Total On-Street Cycling Infrastructure (without Local Street Bikeways)")))
	) + scale_y_continuous(
		label = scales::label_number(suffix = " km")
	) + scale_fill_discrete(
		breaks = rev(c("Vancouver", "Calgary", "Toronto"))
	) + scale_colour_manual(
		values = c("#546ca9", "#c5a43d", "#719d71")
	) + scale_x_continuous(
		breaks = seq(
			settings$year_min + 1,
			settings$year_max,
			by = 1
		),
		limits = c(
			settings$year_min + 1,
			settings$year_max
		)
	) + theme(
		axis.title.y = element_blank(),
		axis.title.x = element_blank(),
		legend.title = element_blank()
	)
	
	# Create cyc tracks infra line plot
	p[[3]] <- pdata %>% filter(
		type == "PBL"
	) %>% group_by(year, city) %>% summarize(
		adj_len_norm = sum(adj_len_norm, na.rm = TRUE)
	) %>% group_by(city) %>% arrange(year) %>% mutate(
		change = adj_len_norm - lag(adj_len_norm)
	) %>% ggplot(aes(
		x = year,
		y = change,
		color = factor(city, levels = c(
			"Vancouver", "Calgary", "Toronto"
		))
	)) + geom_line(
		size = 0.75
	) + geom_point() + geom_vline(
		xintercept = 2019,
		linetype = "dashed",
		color = "gray25",
		size = 0.5
	) + ggtitle(
		bquote(underline(.("Cycle Tracks")))
	) + scale_y_continuous(
		label = scales::label_number(suffix = " km")
	) + scale_colour_manual(
		values = c("#546ca9", "#c5a43d", "#719d71")
	) + scale_x_continuous(
		breaks = seq(
			settings$year_min + 1,
			settings$year_max,
			by = 1
		),
		limits = c(
			settings$year_min + 1,
			settings$year_max
		)
	) + theme(
		axis.title.y = element_blank(),
		axis.title.x = element_blank(),
		legend.title = element_blank()
	)
	
	# Create painted lanes infra line plot
	p[[4]] <- pdata %>% filter(
		type %in% c("PL", "BUF")
	) %>% group_by(year, city) %>% summarize(
		adj_len_norm = sum(adj_len_norm, na.rm = TRUE)
	) %>% group_by(city) %>% arrange(year) %>% mutate(
		change = adj_len_norm - lag(adj_len_norm)
	) %>% ggplot(aes(
		x = year,
		y = change,
		color = factor(city, levels = c(
			"Vancouver", "Calgary", "Toronto"
		))
	)) + geom_line(
		size = 0.75
	) + geom_point() + geom_vline(
		xintercept = 2019,
		linetype = "dashed",
		color = "gray25",
		size = 0.5
	) + geom_hline(
		yintercept = 0,
		color = "gray20",
		size = 0.5
	) + ggtitle(
		bquote(underline(.("Painted and Buffered Lanes")))
	) + scale_y_continuous(
		label = scales::label_number(suffix = " km")
	) + scale_colour_manual(
		values = c("#546ca9", "#c5a43d", "#719d71")
	) + scale_x_continuous(
		breaks = seq(
			settings$year_min + 1,
			settings$year_max,
			by = 1
		),
		limits = c(
			settings$year_min + 1,
			settings$year_max
		)
	) + theme(
		axis.title.y = element_blank(),
		axis.title.x = element_blank(),
		legend.title = element_blank()
	)
	
	# Y-axis title
	y_title <- ggplot() +
		annotate(
			geom = "text",
			x = 1,
			y = 1,
			label = len_title,
			angle = 90,
			size = 5
		) +
		coord_cartesian(clip = "off")+
		theme_void()
	
	# Combine all infra plots together
	out <- (y_title | wrap_plots(p, nrow = length(p))) +
		plot_annotation(
			title = "Yearly Net Change in Cycling Infrastructure\n(per 1000 centreline-km of roadway)",
			caption = sprintf("Years (%s-%s)", settings$year_min + 1, settings$year_max),
			theme = theme(
				plot.title = element_text(hjust = 0.5, size = 16),
				plot.caption = element_text(hjust = 0.5, size = 14)
			)
		) +
		plot_layout(widths = c(0.05, 1))
	return(out)
}
```

# Data

Load raw data provided by Konrad Samsel.

## Vancouver Raw Data {.tabset}

```{r}
vanc_raw <- read_sf("../data/raw/vancouver/Vancouver AS KS Mar26.shp")
```

### Map

**Note**: Only segments with verified installations are shown (n = `r vanc_raw %>% filter(!is.na(INST_TMIN)) %>% nrow` of `r vanc_raw %>% nrow`).

```{r out.width = "100%"}
tmap_mode("view")
tm_shape(vanc_raw %>% filter(!is.na(INST_TMIN))) +
	tm_lines(col = "INST_TMIN", popup.vars = TRUE)
```

### Data

```{r}
vanc_raw %>% as_tibble
```

### Details

```{r}
print(vanc_raw)
```

### Files

Save [vancouver-bikeways-raw-v1.geojson](https://github.com/rrwen/recovr-infracycle/blob/main/data/vancouver-bikeways-raw-v1.geojson) file:

```{r}

# Save geojson
vanc_raw %>%
	write_sf("../data/vancouver-bikeways-raw-v1.geojson", delete_dsn = TRUE)
```

Save [vancouver-bikeways-raw-v1.csv](https://github.com/rrwen/recovr-infracycle/blob/main/data/vancouver-bikeways-raw-v1.csv) file:

```{r}

# Save csv
# st_read("../data/vancouver-bikeways-raw-v1.csv", options = "GEOM_POSSIBLE_NAMES=geometry", crs = "urn:ogc:def:crs:OGC:1.3:CRS84")
vanc_raw %>%
	mutate(
		geometry = st_as_text(geometry),
		geometry_crs = st_crs(vanc_raw)$proj4string,
		.before = geometry
	) %>%
	write_csv("../data/vancouver-bikeways-raw-v1.csv", na = "")
```

## Calgary Raw Data {.tabset}

```{r}
calg_raw <- read_sf("../data/raw/calgary/Calgary Export.shp")
```

### Map

**Note**: Only segments with verified installations are shown (n = `calg_raw %>% filter(!is.na(INST_TMIN)) %>% nrow` of `calg_raw %>% nrow`).

```{r out.width = "100%"}
tmap_mode("view")
tm_shape(calg_raw %>% filter(!is.na(INST_TMIN))) +
	tm_lines(col = "INST_TMIN", popup.vars = TRUE)
```

### Data

```{r}
calg_raw %>% as_tibble
```

### Details

```{r}
print(calg_raw)
```

### Files

Save [calgary-bikeways-raw-v1.geojson](https://github.com/rrwen/recovr-infracycle/blob/main/data/calgary-bikeways-raw-v1.geojson) file:

```{r}

# Save geojson
calg_raw %>%
	write_sf("../data/calgary-bikeways-raw-v1.geojson", delete_dsn = TRUE)
```

Save [calgary-bikeways-raw-v1.csv](https://github.com/rrwen/recovr-infracycle/blob/main/data/calgary-bikeways-raw-v1.csv) file:

```{r}

# Save csv
#st_read("../data/calgary-bikeways-raw-v1.csv", options = "GEOM_POSSIBLE_NAMES=geometry", crs = "urn:ogc:def:crs:OGC:1.3:CRS84")
calg_raw %>%
	mutate(
		geometry = st_as_text(geometry),
		geometry_crs = st_crs(calg_raw)$proj4string,
		.before = geometry
	) %>%
	write_csv("../data/calgary-bikeways-raw-v1.csv", na = "")
```

## Toronto Raw Data {.tabset}

```{r}
toron_raw <- read_sf("../data/raw/toronto/Toronto AS 1323 V3.shp")
```

### Map

**Note**: Only segments with verified installations are shown (n = `r toron_raw %>% filter(!is.na(INST_TMIN)) %>% nrow` of `r toron_raw %>% nrow`).

```{r out.width = "100%"}
tmap_mode("view")
tm_shape(toron_raw %>% filter(!is.na(INST_TMIN))) +
	tm_lines(col = "INST_TMIN", popup.vars = TRUE)
```

### Data

```{r}
toron_raw %>% as_tibble
```

### Details

```{r}
print(toron_raw)
```

### Files

Save [toronto-bikeways-raw-v1.geojson](https://github.com/rrwen/recovr-infracycle/blob/main/data/toronto-bikeways-raw-v1.geojson) file:

```{r}

# Save geojson
toron_raw %>%
	write_sf("../data/toronto-bikeways-raw-v1.geojson", delete_dsn = TRUE)
```

Save [toronto-bikeways-raw-v1.csv](https://github.com/rrwen/recovr-infracycle/blob/main/data/toronto-bikeways-raw-v1.csv) file:

```{r}

# Save csv
# st_read("../data/toronto-bikeways-raw-v1.csv", options = "GEOM_POSSIBLE_NAMES=geometry", crs = "urn:ogc:def:crs:OGC:1.3:CRS84")
toron_raw %>%
	mutate(
		geometry = st_as_text(geometry),
		geometry_crs = st_crs(toron_raw)$proj4string,
		.before = geometry
	) %>%
	write_csv("../data/toronto-bikeways-raw-v1.csv", na = "")
```

# Preprocessing

## Vancouver Preprocessed Data {.tabset}

```{r warning = FALSE}

# Preprocess data
vanc_preprocess <- vanc_raw %>%
	select( # select and rename
		id = object_id,
		street = street_nam,
		status = status,
		road_type = street_seg,
		install_year = year_of_co,
		install_type = bikeway_ty,
		verify_install_year = INST_YR,
		verify_install_date = INST_DATE,
		verify_install_type = INST_TMIN,
		verify_install_comment = INST_COMM,
		verify_upgrade1_year = UPGR1_YR,
		verify_upgrade1_date = UPGR1_DATE,
		verify_upgrade1_type = UPGR1_TMIN,
		verify_upgrade1_comment = UPGR1_COMM,
		verify_upgrade2_year = UPGR2_YR,
		verify_upgrade2_date = UPGR2_DATE,
		verify_upgrade2_type = UPGR2_TMIN,
		verify_upgrade2_comment = UPGR2_COMM,
		verify_misclass = EXCL_REAS
	) %>%
	mutate( # data types
		id = as.character(id),
		street = as.character(street),
		road_type = as.character(road_type),
		install_year = as.numeric(install_year),
		install_type = as.character(install_type),
		verify_install_year = as.numeric(verify_install_year),
		verify_install_date = as.character(verify_install_date),
		verify_install_type = as.character(verify_install_type),
		verify_install_comment = as.character(verify_install_comment),
		verify_upgrade1_year = as.numeric(verify_upgrade1_year),
		verify_upgrade1_date = as.character(verify_upgrade1_date),
		verify_upgrade1_type = as.character(verify_upgrade1_type),
		verify_upgrade1_comment = as.character(verify_upgrade1_comment),
		verify_upgrade2_year = as.numeric(verify_upgrade2_year),
		verify_upgrade2_date = as.character(verify_upgrade2_date),
		verify_upgrade2_type = as.character(verify_upgrade2_type),
		verify_upgrade2_comment = as.character(verify_upgrade2_comment),
		verify_misclass = as.character(verify_misclass)
	) %>%
	mutate( # clean values
		install_year = na_if(install_year, 0),
		verify_install_year = na_if(verify_install_year, 0),
		verify_install_date = na_if(verify_install_date, "NA"),
		verify_install_type = na_if(verify_install_type, "NA") %>%
			str_replace_all("[^[:alpha:]]|\\s", ""),
		verify_install_comment = na_if(verify_install_comment, "NA"),
		verify_upgrade1_year = na_if(verify_upgrade1_year, 0),
		verify_upgrade1_date = na_if(verify_upgrade1_date, "NA"),
		verify_upgrade1_type = na_if(verify_upgrade1_type, "NA") %>%
			str_replace_all("[^[:alpha:]]|\\s", ""),
		verify_upgrade1_comment = na_if(verify_upgrade1_comment, "NA"),
		verify_upgrade2_year = na_if(verify_upgrade2_year, 0),
		verify_upgrade2_date = na_if(verify_upgrade2_date, "NA"),
		verify_upgrade2_type = na_if(verify_upgrade2_type, "NA") %>%
			str_replace_all("[^[:alpha:]]|\\s", ""),
		verify_upgrade2_comment = na_if(verify_upgrade2_comment, "NA"),
		verify_misclass =  na_if(verify_misclass, "NA") %>%
			str_trim %>%
			str_to_title
	) %>%
	mutate( # add column for non-verified infra types
		no_verify_install_type = if_else(
			is.na(verify_install_type) & install_type == "Local Street",
			"Local Street",
			NA
		),
		.after = verify_misclass
	) %>%
	mutate( # add local street as non-verified LSB
		verify_install_type = if_else(
			is.na(verify_install_type) & install_type == "Local Street",
			"LSB",
			verify_install_type
		),
		verify_install_year = if_else(
			is.na(verify_install_year) & install_type == "Local Street",
			install_year,
			verify_install_year
		)
	) %>%
	mutate( # create col for recoded road types
		road_type_recode = case_when( # create road types
	        road_type %in% c( # arterial equiv
	        	"Arterial"
	        ) ~ "Arterial",
	        road_type %in% c( # collector equiv
	        	"Collector",
	        	"Secondary Arterial",
	        	"Sec Arterial"
	        ) ~ "Collector",
	        road_type %in% c( # local equiv
	        	"Lane",
	        	"Residential",
	        	"Leased",
	        	"Recreational"
	        ) ~ "Local",
	        .default = road_type
	    ),
		.after = road_type
	) %>%
	mutate( # create col for segment lengths in km
		geometry_len_km = as.numeric(st_length(geometry)) / 1000,
		.before = geometry
	)
```

### Map

**Note**: Only the first 1000 records are shown as a sample.

```{r out.width = "100%"}
tmap_mode("view")
tm_shape(
	vanc_preprocess %>% head(1000)) +
	tm_lines(col = "verify_install_type", popup.vars = TRUE)
```

### Data

```{r}
vanc_preprocess %>% as_tibble
```

### Details

```{r}
print(vanc_preprocess)
```

### Files

Save [vancouver-bikeways-preprocess-v1.geojson](https://github.com/rrwen/recovr-infracycle/blob/main/data/vancouver-bikeways-preprocess-v1.geojson) file:

```{r}

# Save geojson
vanc_preprocess %>%
	write_sf("../data/vancouver-bikeways-preprocess-v1.geojson", delete_dsn = TRUE)
```

Save [vancouver-bikeways-preprocess-v1.csv](https://github.com/rrwen/recovr-infracycle/blob/main/data/vancouver-bikeways-preprocess-v1.csv) file:

```{r}

# Save csv
# st_read("../data/vancouver-bikeways-preprocess-v1.csv", options = "GEOM_POSSIBLE_NAMES=geometry", crs = "urn:ogc:def:crs:OGC:1.3:CRS84")
vanc_preprocess %>%
	mutate(
		geometry = st_as_text(geometry),
		geometry_crs = st_crs(vanc_preprocess)$proj4string,
		.before = geometry
	) %>%
	write_csv("../data/vancouver-bikeways-preprocess-v1.csv", na = "")
```

## Calgary Preprocessed Data {.tabset}

```{r warning = FALSE}

# Preprocess data
calg_preprocess <- calg_raw %>%
	select( # select and rename
		id = shape_id,
		street = STREET,
		status = STATUS,
		road_type = CENTL_CLAS,
		install_year = date_creat,
		install_type = BICYCLE_CL,
		verify_install_year = INST_YR,
		verify_install_date = INST_DATE,
		verify_install_type = INST_TMIN,
		verify_install_comment = INST_COMM,
		verify_upgrade1_year = UPGR1_YR,
		verify_upgrade1_date = UPGR1_DATE,
		verify_upgrade1_type = UPGR1_TMIN,
		verify_upgrade1_comment = UPGR1_COMM,
		verify_upgrade2_year = UPGR2_YR,
		verify_upgrade2_date = UPGR2_DATE,
		verify_upgrade2_type = UPGR2_TMIN,
		verify_upgrade2_comment = UPGR2_COMM,
		verify_misclass = EXCL_REAS
	) %>%
	mutate( # data types
		id = as.character(id),
		street = as.character(street),
		road_type = as.character(road_type),
		install_year = as.numeric(year(install_year)),
		install_type = as.character(install_type),
		verify_install_year = as.numeric(verify_install_year),
		verify_install_date = as.character(verify_install_date),
		verify_install_type = as.character(verify_install_type),
		verify_install_comment = as.character(verify_install_comment),
		verify_upgrade1_year = as.numeric(verify_upgrade1_year),
		verify_upgrade1_date = as.character(verify_upgrade1_date),
		verify_upgrade1_type = as.character(verify_upgrade1_type),
		verify_upgrade1_comment = as.character(verify_upgrade1_comment),
		verify_upgrade2_year = as.numeric(verify_upgrade2_year),
		verify_upgrade2_date = as.character(verify_upgrade2_date),
		verify_upgrade2_type = as.character(verify_upgrade2_type),
		verify_upgrade2_comment = as.character(verify_upgrade2_comment),
		verify_misclass = as.character(verify_misclass)
	) %>%
	mutate( # clean values
		install_year = na_if(install_year, 0),
		verify_install_year = na_if(verify_install_year, 0),
		verify_install_date = na_if(verify_install_date, "NA"),
		verify_install_type = na_if(verify_install_type, "NA") %>%
			str_replace_all("[^[:alpha:]]|\\s", ""),
		verify_install_comment = na_if(verify_install_comment, "NA"),
		verify_upgrade1_year = na_if(verify_upgrade1_year, 0),
		verify_upgrade1_date = na_if(verify_upgrade1_date, "NA"),
		verify_upgrade1_type = na_if(verify_upgrade1_type, "NA") %>%
			str_replace_all("[^[:alpha:]]|\\s", ""),
		verify_upgrade1_comment = na_if(verify_upgrade1_comment, "NA"),
		verify_upgrade2_year = na_if(verify_upgrade2_year, 0),
		verify_upgrade2_date = na_if(verify_upgrade2_date, "NA"),
		verify_upgrade2_type = na_if(verify_upgrade2_type, "NA") %>%
			str_replace_all("[^[:alpha:]]|\\s", ""),
		verify_upgrade2_comment = na_if(verify_upgrade2_comment, "NA"),
		verify_misclass =  na_if(verify_misclass, "NA") %>%
			str_trim %>%
			str_to_title
	) %>%
	mutate( # create col for recoded road types
		road_type_recode = case_when( # create road types
	        road_type %in% c( # arterial equiv
	        	 "Arterial Street",
	        	 "Industrial Arterial",
	        	 "Local Arterial",
	        	 "Parkway",
	        	 "Urban Boulevard"
	        ) ~ "Arterial",
	        road_type %in% c( # collector equiv
	        	"Neighbourhood Boulevard",
	        	"Collector",
	        	"Primary Collector"
	        ) ~ "Collector",
	        road_type %in% c( # local equiv
	        	"Access Route",
	        	"Residential Street",
	        	"Activity Center Street",
	        	"Historic Road Allowance",
	        	"Lanes (Alleys)",
	        	"Industrial Street"
	        ) ~ "Local",
	        .default = road_type
	    ),
		.after = road_type
	) %>%
	mutate( # create col for segment lengths in km
		geometry_len_km = as.numeric(st_length(geometry)) / 1000,
		.before = geometry
	) %>%
	st_transform(4326) # reproject to WGS84
```

### Map

**Note**: Only the first 1000 records are shown as a sample.

```{r out.width = "100%"}
tmap_mode("view")
tm_shape(calg_preprocess %>% head(1000)) +
	tm_lines(col = "verify_install_type", popup.vars = TRUE)
```

### Data

```{r}
calg_preprocess %>% as_tibble
```

### Details

```{r}
print(calg_preprocess)
```

### Files

Save [calgary-bikeways-preprocess-v1.geojson](https://github.com/rrwen/recovr-infracycle/blob/main/data/calgary-bikeways-preprocess-v1.geojson) file:

```{r}

# Save geojson
calg_preprocess %>%
	write_sf("../data/calgary-bikeways-preprocess-v1.geojson", delete_dsn = TRUE)
```

Save [calgary-bikeways-preprocess-v1.csv](https://github.com/rrwen/recovr-infracycle/blob/main/data/calgary-bikeways-preprocess-v1.csv) file:

```{r}

# Save csv
# st_read("../data/calgary-bikeways-preprocess-v1.csv", options = "GEOM_POSSIBLE_NAMES=geometry", crs = "urn:ogc:def:crs:OGC:1.3:CRS84")
calg_preprocess %>%
	mutate(
		geometry = st_as_text(geometry),
		geometry_crs = st_crs(calg_preprocess)$proj4string,
		.before = geometry
	) %>%
	write_csv("../data/calgary-bikeways-preprocess-v1.csv", na = "")
```

## Toronto Preprocessed Data {.tabset}

```{r warning = FALSE}

# Preprocess data
toron_preprocess <- toron_raw %>%
	select( # select and rename
		id = OBJECTI2,
		street = STREET_7,
		street_from = FROM_ST8,
		street_to = TO_STRE9,
		road_type = M_FEATUR36,
		install_year = C_INST_YR,
		install_type = INFRA_H20,
		verify_install_year = INST_YR,
		verify_install_date = INST_DATE,
		verify_install_type = INST_TMIN,
		verify_install_comment = INST_COMM,
		verify_upgrade1_year = UPGR1_YR,
		verify_upgrade1_date = UPGR1_DATE,
		verify_upgrade1_type = UPGR1_TMIN,
		verify_upgrade1_comment = UPGR1_COMM,
		verify_upgrade2_year = UPGR2_YR,
		verify_upgrade2_date = UPGR2_DATE,
		verify_upgrade2_type = UPGR2_TMIN,
		verify_upgrade2_comment = UPGR2_COMM,
		verify_misclass = EXCL_REAS
	) %>%
	mutate( # data types
		id = as.character(id),
		street = as.character(street),
		street_from = as.character(street_from),
		street_to = as.character(street_to),
		road_type = as.character(road_type),
		install_year = as.numeric(install_year),
		install_type = as.character(install_type),
		verify_install_year = as.numeric(verify_install_year),
		verify_install_date = as.character(verify_install_date),
		verify_install_type = as.character(verify_install_type),
		verify_install_comment = as.character(verify_install_comment),
		verify_upgrade1_year = as.numeric(verify_upgrade1_year),
		verify_upgrade1_date = as.character(verify_upgrade1_date),
		verify_upgrade1_type = as.character(verify_upgrade1_type),
		verify_upgrade1_comment = as.character(verify_upgrade1_comment),
		verify_upgrade2_year = as.numeric(verify_upgrade2_year),
		verify_upgrade2_date = as.character(verify_upgrade2_date),
		verify_upgrade2_type = as.character(verify_upgrade2_type),
		verify_upgrade2_comment = as.character(verify_upgrade2_comment),
		verify_misclass = as.character(verify_misclass)
	) %>%
	mutate( # clean values
		install_year = na_if(install_year, 0),
		verify_install_year = na_if(verify_install_year, 0),
		verify_install_date = na_if(verify_install_date, "NA"),
		verify_install_type = na_if(verify_install_type, "NA") %>%
			str_replace_all("[^[:alpha:]]|\\s", ""),
		verify_install_comment = na_if(verify_install_comment, "NA"),
		verify_upgrade1_year = na_if(verify_upgrade1_year, 0),
		verify_upgrade1_date = na_if(verify_upgrade1_date, "NA"),
		verify_upgrade1_type = na_if(verify_upgrade1_type, "NA") %>%
			str_replace_all("[^[:alpha:]]|\\s", ""),
		verify_upgrade1_comment = na_if(verify_upgrade1_comment, "NA"),
		verify_upgrade2_year = na_if(verify_upgrade2_year, 0),
		verify_upgrade2_date = na_if(verify_upgrade2_date, "NA"),
		verify_upgrade2_type = na_if(verify_upgrade2_type, "NA") %>%
			str_replace_all("[^[:alpha:]]|\\s", ""),
		verify_upgrade2_comment = na_if(verify_upgrade2_comment, "NA"),
		verify_misclass =  na_if(verify_misclass, "NA") %>%
			str_trim %>%
			str_to_title
	) %>%
	mutate( # create col for recoded road types
		road_type_recode = case_when( # create road types
	        road_type %in% c( # arterial equiv
	        	"Major Arterial",
	        	"Major Arterial Ramp",
	        	"Minor Arterial"
	        ) ~ "Arterial",
	        road_type %in% c( # collector equiv
	        	"Collector"
	        ) ~ "Collector",
	        road_type %in% c(  # local equiv
	        	"Local",
	        	"Other"
	        ) ~ "Local",
	        .default = road_type
	    ),
		.after = road_type
	) %>%
	mutate( # create col for segment lengths in km
		geometry_len_km = as.numeric(st_length(geometry)) / 1000,
		.before = geometry
	)
```

### Map

**Note**: Only the first 1000 records are shown as a sample.

```{r out.width = "100%"}
tmap_mode("view")
tm_shape(toron_preprocess %>% head(1000)) +
	tm_lines(col = "verify_install_type", popup.vars = TRUE)
```

### Data

```{r}
toron_preprocess %>% as_tibble
```

### Details

```{r}
print(toron_preprocess)
```

### Files

Save [toronto-bikeways-preprocess-v1.geojson](https://github.com/rrwen/recovr-infracycle/blob/main/data/toronto-bikeways-preprocess-v1.geojson) file:

```{r}

# Save geojson
toron_preprocess %>%
	write_sf("../data/toronto-bikeways-preprocess-v1.geojson", delete_dsn = TRUE)
```

Save [toronto-bikeways-preprocess-v1.csv](https://github.com/rrwen/recovr-infracycle/blob/main/data/toronto-bikeways-preprocess-v1.csv) file:

```{r}

# Save csv
# st_read("../data/toronto-bikeways-preprocess-v1.csv", options = "GEOM_POSSIBLE_NAMES=geometry", crs = "urn:ogc:def:crs:OGC:1.3:CRS84")
toron_preprocess %>%
	mutate(
		geometry = st_as_text(geometry),
		geometry_crs = st_crs(toron_preprocess)$proj4string,
		.before = geometry
	) %>%
	write_csv("../data/toronto-bikeways-preprocess-v1.csv", na = "")
```

# Inclusion and Exclusion Criteria

Apply filters for inclusion and exclusion criteria using function `filter_criteria` as described in the methods and Appendix 2.

```{r}

# Build filter criteria
cities_criteria <- list(
	vancouver = list(
		city = "vancouver",
		data = vanc_preprocess,
		data_date = "January 2023",
		data_url = "https://opendata.vancouver.ca/explore/dataset/bikeways/information",
		infra_col = "install_type",
		infra_filter = c("Painted Lanes", "Protected Bike Lanes", "Local Street"),
		road_col = "road_type",
		road_filter = c("Off-street"),
		geom_col = "geometry",
		geom_unit = "km",
		geom_filter = TRUE,
		misclass_col = "verify_misclass",
		misclass_filter = c(NA, "NA"),
		noverify_col = "no_verify_install_type",
		noverify_filter = c("Local Street")
	),
	calgary = list(
		city = "calgary",
		data = calg_preprocess,
		data_date = "January 2023",
		data_url = "https://data.calgary.ca/Transportation-Transit/Calgary-Bikeways/jjqk-9b73",
		infra_col = "install_type",
		infra_filter = c("Bicycle Lane", "Cycle Track"),
		status_col = "status",
		status_filter = c("INACTIVE", "PLANNED"),
		geom_col = "geometry",
		geom_unit = "km",
		geom_filter = TRUE,
		misclass_col = "verify_misclass",
		misclass_filter = c(NA, "NA")
	),
	toronto = list(
		city = "toronto",
		data = toron_preprocess,
		data_date = "January 2023",
		data_url = "https://open.toronto.ca/dataset/cycling-network",
		infra_col = "install_type",
		infra_filter = c("Bi-Directional Cycle Track", "Bike Lane", "Bike Lane - Buffered", "Bike Lane - Contraflow", "Cycle Track", "Cycle Track - Contraflow"),
		geom_col = "geometry",
		geom_unit = "km",
		geom_filter = TRUE,
		misclass_col = "verify_misclass",
		misclass_filter = c(NA, "NA")
	)
)

# Apply filter criteria for all cities
criteria_data <- filter_criteria(cities_criteria)
```

## Vancouver Filtered Data {.tabset}

```{r}
vanc <- criteria_data$vancouver$data_filter
```

### Map

**Note**: Only the first 1000 records are shown as a sample.

```{r out.width = "100%"}
tmap_mode("view")
tm_shape(vanc %>% head(1000)) +
	tm_lines(col = "verify_install_type", popup.vars = TRUE)
```

### Data

```{r}
vanc %>% as_tibble
```

### Details

```{r}
print(vanc)
```

## Calgary Filtered Data {.tabset}

```{r}
calg<- criteria_data$calgary$data_filter
```

### Map

**Note**: Only the first 1000 records are shown as a sample.

```{r out.width = "100%"}
tmap_mode("view")
tm_shape(calg %>% head(1000)) +
	tm_lines(col = "verify_install_type", popup.vars = TRUE)
```

### Data

```{r}
calg %>% as_tibble
```

### Details

```{r}
print(calg)
```

## Toronto Filtered Data {.tabset}

```{r}
toron <- criteria_data$toronto$data_filter
```

### Map

**Note**: Only the first 1000 records are shown as a sample.

```{r out.width = "100%"}
tmap_mode("view")
tm_shape(toron %>% head(1000)) +
	tm_lines(col = "verify_install_type", popup.vars = TRUE)
```

### Data

```{r}
toron %>% as_tibble
```

### Details

```{r}
print(toron)
```

# Map Data Preparation

Prepare data and settings for figures with maps.

```{r}
map_data <- list(
	vancouver = list(
		title = "Vancouver, CA",
		data = vanc,
		downtown_bbox = c(
			xmin = -123.143450,
			ymin = 49.269529,
			xmax = -123.095584,
			ymax = 49.296229
		)
	),
	calgary = list(
		title = "Calgary, CA",
		data = calg,
		downtown_bbox = c(
			xmin = -114.127909,
			ymin = 51.006626,
			xmax = -113.975817,
			ymax = 51.081312
		)
	),
	toronto = list(
		title = "Toronto, CA",
		data = toron,
		downtown_bbox = c(
			xmin = -79.300395,
			ymin = 43.636621,
			xmax = -79.489565,
			ymax = 43.698150
		)
	)
)
```

# Figures

## Figure 1: Flow diagram of inclusion criteria for bikeway segments in Vancouver, Calgary, and Toronto. {.tabset}

This flowchart provides a high-level overview of the segment inclusions and exclusions for each municipality. Data from Calgary were specific to on-street routes only. For detailed flow diagrams specific to each municipality, please refer to the Appendix.

```{r}
fig1 <- list()
fig1$plot <- diag_criteria(criteria_data)
```

### Figure

```{r}
fig1$plot
```

### Files

```{r}      

# Save pdf
fig1$plot %>%
    export_svg %>%
    charToRaw %>%
    rsvg_pdf("../manuscript/figures/fig-methods.pdf")

# Save png
fig1$plot %>%
    export_svg %>%
    charToRaw %>%
    rsvg_png("../manuscript/figures/fig-methods.png")
```

* [manuscript/figures/fig-methods.pdf](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/fig-methods.pdf)
* [manuscript/figures/fig-methods.png](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/fig-methods.png)

## Figure 2: Changes in dedicated cycling infrastructure between 2009 and 2022 for Vancouver, Calgary, and Toronto by infrastructure category. {.tabset}

Assessed using roadway centreline-km, with infrastructure classifications determined by the most protective element present along each road segment.

```{r}
fig2 <- list()
fig2$plot <- plot_yearly_len_infra(
	df_list = list(
		"Vancouver, CA (~2223.7 km Total Roadway)" = list(
			data = vanc,
			roadway_per = 1000,
			roadway_total = 2223.7
		),
		"Vancouver, CA (without Local Street Bikeways)" = list(
			data = vanc %>% filter(verify_install_type != "LSB"),
			roadway_per = 1000,
			roadway_total = 2223.7
		),
		"Calgary, CA (~7931.2 km Total Roadway)" = list(
			data = calg,
			roadway_per = 1000,
			roadway_total = 7931.2
		),
		"Toronto, CA (~5579.4 km Total Roadway)" = list(
			data = toron,
			roadway_per = 1000,
			roadway_total = 5579.4
		)
	),
	len_title = "Length per 1000 Centreline km of Total Roadway",
	len_per = TRUE,
	line_km = 10
)
```

### Figure

```{r fig.width = 8.5, fig.height = 12.5}
fig2$plot
```

### Files

```{r}

# Save pdf
ggsave(
	"../manuscript/figures/fig-yearly-len.pdf",
	fig2$plot,
	width = 8.5,
	height = 12.5
)

# Save png
ggsave(
	"../manuscript/figures/fig-yearly-len.png",
	fig2$plot,
	width = 8.5,
	height = 12.5
)
```

* [manuscript/figures/fig-yearly-len.pdf](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/fig-yearly-len.pdf)
* [manuscript/figures/fig-yearly-len.png](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/fig-yearly-len.png)

## Figure 3: Yearly net change in cycle route infrastructure by municipality, standardized per 1000 centerline-km of roadway. {.tabset}

The net change considers both the installation of new facilities, and the removal of existing infrastructure, such as when an existing facility is upgraded. Cycle route infrastructure is defined by the most protective element along a street centreline. This reflects the overall modifications made within each municipality over the course of the study period (2009-2022).

```{r}
fig3 <- list()
fig3$plot <- plot_yearly_change(
	df_list = list(
		"Vancouver, CA" = list(
			city = "Vancouver",
			data = vanc,
			roadway_per = 1000,
			roadway_total = 2223.7
		),
		"Calgary, CA" = list(
			city = "Calgary",
			data = calg,
			roadway_per = 1000,
			roadway_total = 7931.2
		),
		"Toronto, CA" = list(
			city = "Toronto",
			data = toron,
			roadway_per = 1000,
			roadway_total = 5579.4
		)
	),
	len_title = "Length per 1000 Centreline km of Total Roadway"
)
```

### Figure

```{r fig.width = 8, fig.height = 12}
fig3$plot
```

### Files

```{r}

# Save pdf
ggsave(
	"../manuscript/figures/fig-yearly-change.pdf",
	fig3$plot,
	width = 8,
	height = 12
)

# Save png
ggsave(
	"../manuscript/figures/fig-yearly-change.png",
	fig3$plot,
	width = 8,
	height = 12
)
```

* [manuscript/figures/fig-yearly-change.pdf](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/fig-yearly-change.pdf)
* [manuscript/figures/fig-yearly-change.png](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/fig-yearly-change.png)

## Figure 4: Changes in Dedicated On-Street Infrastructure Since January 2020 for Vancouver, Calgary, and Toronto. {.tabset}

New installations of dedicated infrastructure are denoted in green, upgrades from a previous dedicated infrastructure type are denoted in orange.  Basemap from OpenStreetMap and Carto (Positron).


```{r}
fig4 <- list()
fig4$plot <- map_infra(map_data, scale_prop = 0.35)
```

### Figure

```{r fig.width=9, fig.height=11}
fig4$plot %>% print
```

### Files

```{r}

# Save pdf
ggsave(
	"../manuscript/figures/fig-maps.pdf",
	fig4$plot,
	width = 8,
	height = 12
)

# Save png
ggsave(
	"../manuscript/figures/fig-maps.png",
	fig4$plot,
	width = 8,
	height = 12
)
```

* [manuscript/figures/fig-maps.pdf](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/fig-maps.pdf)
* [manuscript/figures/fig-maps.png](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/fig-maps.png)

# Appendix 1 - Supplementary Results

## Supplementary Figure 1: Enlarged Map. Changes in Dedicated On-Street Infrastructure Between 2020-2021 for the Municipality of Vancouver, CA. {.tabset}

New installations of dedicated infrastructure are denoted in green, upgrades from a previous dedicated infrastructure type are denoted in orange. Basemap from OpenStreetMap and Carto (Positron).

```{r}
sfig1 <- list()
sfig1$plot <- map_infra_detail(
    map_data,
    "vancouver",
    map_inset_position = c(
        left = -0.9,
        bottom = 0.65,
        right = 1.2125,
        top = 0.99
    ),
    map_ratio = 1.5,
    map_inset_ratio = 1.2
)
```

### Figure

```{r fig.width = 8, fig.height = 6}
sfig1$plot %>% print
```

### Files

```{r}

# Save pdf
ggsave(
	"../manuscript/figures/sfig-map-vanc.pdf",
	sfig1$plot,
	width = 8,
	height = 6
)

# Save png
ggsave(
	"../manuscript/figures/sfig-map-vanc.png",
	sfig1$plot,
	width = 8,
	height = 6
)
```

* [manuscript/figures/sfig-map-vanc.pdf](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/sfig-map-vanc.pdf)
* [manuscript/figures/sfig-map-vanc.png](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/sfig-map-vanc.png)

## Supplementary Figure 2: Enlarged Map. Changes in Dedicated On-Street Infrastructure Between 2020-2022 for the Municipality of Calgary, CA. {.tabset}

New installations of dedicated infrastructure are denoted in green, upgrades of dedicated infrastructure are denoted in orange. Basemap from OpenStreetMap and Carto (Positron).

```{r}
sfig2 <- list()
sfig2$plot <- map_infra_detail(
    map_data,
    "calgary",
    map_inset_position = c(
        left = -0.85,
        bottom = 0.65,
        right = 1.2125,
        top = 0.99
    ),
    map_ratio = 1.25,
    map_inset_ratio = 1.2
)
```

### Figure

```{r fig.width = 8, fig.height = 6}
sfig2$plot %>% print
```

### Files

```{r}

# Save pdf
ggsave(
	"../manuscript/figures/sfig-map-calg.pdf",
	sfig2$plot,
	width = 8,
	height = 6
)

# Save png
ggsave(
	"../manuscript/figures/sfig-map-calg.png",
	sfig2$plot,
	width = 8,
	height = 6
)
```

* [manuscript/figures/sfig-map-calg.pdf](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/sfig-map-calg.pdf)
* [manuscript/figures/sfig-map-calg.png](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/sfig-map-calg.png)

## Supplementary Figure 3: Enlarged Map. Changes in Dedicated On-Street Infrastructure Between 2020-2022 for the Municipality of Toronto, CA. {.tabset}

New installations of dedicated infrastructure are denoted in green, upgrades of dedicated infrastructure are denoted in orange. Basemap from OpenStreetMap and Carto (Positron).

```{r}
sfig3 <- list()
sfig3$plot <- map_infra_detail(
    map_data,
    "toronto",
    map_inset_position = c(
        left = -0.85,
        bottom = 0.65,
        right = 1.38,
        top = 0.99
    ),
    map_ratio = 1.75,
    map_inset_ratio = 2.5
)
```

### Figure

```{r fig.width = 8, fig.height = 6}
sfig3$plot %>% print
```

### Files

```{r}

# Save pdf
ggsave(
	"../manuscript/figures/sfig-map-toron.pdf",
	sfig3$plot,
	width = 8,
	height = 6
)

# Save png
ggsave(
	"../manuscript/figures/sfig-map-toron.png",
	sfig3$plot,
	width = 8,
	height = 6
)
```

* [manuscript/figures/sfig-map-toron.pdf](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/sfig-map-toron.pdf)
* [manuscript/figures/sfig-map-toron.png](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/sfig-map-toron.png)

## Supplementary Table 1: Total Length of Dedicated On-Street Cycling Infrastructure between 2009 and 2022, for Vancouver, Calgary, and Toronto (Canada). {.tabset}

Each entry denotes the aggregated length of infrastructure existing at the conclusion the calendar year. Lengths are measured in roadway centreline-km, with cycling infrastructure classified according to the side of the road featuring the most protective element. Rows noted in light red denote infrastructure changes following the start of the COVID-19 pandemic.

```{r}

# Setup table list
stab1 <- list()

# Calculated adjusted yearly road lengths for each type
stab1$data <- bind_rows(
	calc_yearly_adj_len(vanc) %>% mutate(city = "Vancouver"),
	calc_yearly_adj_len(calg) %>% mutate(city = "Calgary"),
	calc_yearly_adj_len(toron) %>% mutate(city = "Toronto")
) %>%
    pivot_wider( # pivot infra types per col
    	names_from = type,
    	values_from = adj_len,
    	values_fill = 0
    ) %>%
    group_by(year, city) %>%
    summarize( # Calculate yearly road len for each type
        PL = round(sum(PL, na.rm = TRUE), 2),
        BUF = round(sum(BUF, na.rm = TRUE), 2),
        CT = round(sum(PBL, na.rm = TRUE), 2)
    ) %>%
    filter(
        year >= settings$year_min &
        year <= settings$year_max
    ) %>%
    ungroup() %>% mutate( # Calc total road len based on type
        TOTAL = PL + BUF + CT
    ) %>%
    group_by(city) %>% arrange(year) %>% mutate(
        Change = TOTAL - lag(TOTAL) # change in total road len
    ) %>%
	rename(
		Year = year,
		City = city
	)

# Create side by side tables by city using joins
stab1$data <- stab1$data %>% filter(City == "Vancouver") %>%
	select(-City) %>%
	left_join(
		stab1$data %>% filter(City == "Calgary"),
		by = "Year"
	) %>%
	left_join(
		stab1$data %>% filter(City == "Toronto"),
		by = "Year"
	) %>%
	select(
		-City,
		-City.x,
		-City.y
	)

# Gen table
options(knitr.kable.NA = "")
stab1$table <- stab1$data %>%
	kable(
	  col.names = gsub("\\.x|\\.y", "", names(.)),
	  booktabs = T
	) %>%
	kable_classic() %>%
	column_spec(
		c(5,6, 10, 11, 15, 16),
		background = "grey90"
	) %>%
	row_spec(
		0,
		bold = T
	) %>%
	row_spec(
		12:14,
		background = "grey70"
	) %>%
	add_header_above(
		c(" " = 1, "Vancouver" = 5, "Calgary" = 5, "Toronto" = 5),
		align = "left"
	) %>%
	add_header_above(
		c(" " = 1, "Measured by centreline-km of roadway" = 15),
		italic = T,
		bold = F,
		align = "left"
	) %>%
	add_header_above(
		c(" " = 1, "Total Length of Roadways with Dedicated Cycling Infrastructure by Year (2009-2022)" = 15),
		align = "left",
		line = F
	)
```

### Table

```{r}
stab1$table
```

### Files

```{r}
save_kable(stab1$table, "../manuscript/figures/tab-yearly-len.pdf")
save_kable(stab1$table, "../manuscript/figures/tab-yearly-len.png", zoom = 2)
```

* [manuscript/figures/tab-yearly-len.pdf](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/tab-yearly-len.pdf)
* [manuscript/figures/tab-yearly-len.png](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/tab-yearly-len.png)

## Supplementary Figure 4: Changes in dedicated cycling infrastructure between 2009 and 2021 for the Municipality of Vancouver, CA. {.tabset}

By (A) roadway classification, and (B) infrastructure distribution within each road class. Assessed using roadway centreline-km, with infrastructure classification determined by the most protective element present along each road segment.

```{r}
sfig4 <- list()
sfig4$plot <- plot_yearly_len_road(
	vanc,
	title = "Roadways with Dedicated Cycling Infrastructure (Vancouver, CA)"
)
```

### Figure

```{r fig.width=18, fig.height=10}
sfig4$plot
```

### Files

```{r}

# Save pdf
ggsave(
	"../manuscript/figures/sfig-change-vanc.pdf",
	sfig4$plot,
	width = 18,
	height = 10
)

# Save png
ggsave(
	"../manuscript/figures/sfig-change-vanc.png",
	sfig4$plot,
	width = 18,
	height = 10
)
```

* [manuscript/figures/sfig-change-vanc.pdf](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/sfig-change-vanc.pdf)
* [manuscript/figures/sfig-change-vanc.png](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/sfig-change-vanc.png)

## Supplementary Figure 5: Changes in dedicated cycling infrastructure between 2009 and 2022 for the Municipality of Calgary, CA. {.tabset}

By (A) roadway classification, and (B) infrastructure distribution within each road class. Assessed using roadway centreline-km, with infrastructure classification determined by the most protective element present along each road segment.

```{r}
sfig5 <- list()
sfig5$plot <- plot_yearly_len_road(
	calg,
	title = "Roadways with Dedicated Cycling Infrastructure (Calgary, CA)"
)
```

### Figure

```{r fig.width=18, fig.height=10}
sfig5$plot
```

### Files

```{r}

# Save pdf
ggsave(
	"../manuscript/figures/sfig-change-calg.pdf",
	sfig5$plot,
	width = 18,
	height = 10
)

# Save png
ggsave(
	"../manuscript/figures/sfig-change-calg.png",
	sfig5$plot,
	width = 18,
	height = 10
)
```

* [manuscript/figures/sfig-change-calg.pdf](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/sfig-change-calg.pdf)
* [manuscript/figures/sfig-change-calg.png](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/sfig-change-calg.png)

## Supplementary Figure 6: Changes in dedicated cycling infrastructure between 2009 and 2022 for the Municipality of Toronto, CA. {.tabset}

By (A) roadway classification, and (B) infrastructure distribution within each road class. Assessed using roadway centreline-km, with infrastructure classification determined by the most protective element present along each road segment.

```{r}
sfig6 <- list()
sfig6$plot <- plot_yearly_len_road(
	toron,
	title = "Roadways with Dedicated Cycling Infrastructure (Toronto, CA)"
)
```

### Figure

```{r fig.width=18, fig.height=10}
sfig6$plot
```

### Files

```{r}

# Save pdf
ggsave(
	"../manuscript/figures/sfig-change-toron.pdf",
	sfig6$plot,
	width = 18,
	height = 10
)

# Save png
ggsave(
	"../manuscript/figures/sfig-change-toron.png",
	sfig6$plot,
	width = 18,
	height = 10
)
```

* [manuscript/figures/sfig-change-toron.pdf](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/sfig-change-toron.pdf)
* [manuscript/figures/sfig-change-toron.png](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/sfig-change-toron.png)

## Supplementary Figure 7: A comparative analysis between municipal data and verified data on the installation years for cycling infrastructure in Vancouver, CA. {.tabset}

```{r}
# Create the plot
sfig7 <- plot_yearly_diff(
	vanc %>% filter(
		is.na(no_verify_install_type)
	),
	title = "Difference in Installation Years, Comparing City Data and Verified Data: Vancouver, CA",
	out_data = TRUE
)

# Calc metrics for description
sfig7_n <- sum(sfig7$data$n)
sfig7_0 <- sfig7$data %>%  # perc correct
	filter(year_diff == 0) %>%
	pull(n_perc) %>%
	round(1)
sfig7_pm1 <- sfig7$data %>% # perc plus/minus 1
	filter(year_diff >= -1 & year_diff <= 1) %>%
	pull(n_perc) %>%
	sum %>%
	round(1)
```

Any data where a city provided and verified installation years were missing or the verified year occurred earlier or equal to the start of the study period (2009) has been excluded from analysis, yielding n=`r sfig7_n` segments. The graph shows that `r sfig7_0`% of the included segments had the correct installation year as per the city's data, and `r sfig7_pm1`% were accurate within a range of 1 year.

### Figure

```{r fig.width = 8, fig.height = 5}
sfig7$plot
```

### Files

```{r}

# Save pdf
ggsave(
	"../manuscript/figures/sfig-compare-vanc.pdf",
	sfig7$plot,
	width = 18,
	height = 10
)

# Save png
ggsave(
	"../manuscript/figures/sfig-compare-vanc.png",
	sfig7$plot,
	width = 18,
	height = 10
)
```

* [manuscript/figures/sfig-compare-vanc.pdf](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/sfig-compare-vanc.pdf)
* [manuscript/figures/sfig-compare-vanc.png](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/sfig-compare-vanc.png)

## Supplementary Figure 8: A comparative analysis between municipal data and verified data on the installation years for cycling infrastructure in Calgary, CA. {.tabset}

```{r}

# Create the plot
sfig8 <- plot_yearly_diff(
	calg,
	title = "Difference in Installation Years, Comparing City Data and Verified Data: Calgary, CA",
	out_data = TRUE
)

# Calc metrics for description
sfig8_n <- sum(sfig8$data$n)
sfig8_0 <- sfig8$data %>%  # perc correct
	filter(year_diff == 0) %>%
	pull(n_perc) %>%
	round(1)
sfig8_pm1 <- sfig8$data %>% # perc plus/minus 1
	filter(year_diff >= -1 & year_diff <= 1) %>%
	pull(n_perc) %>%
	sum %>%
	round(1)
```

Any data where a city provided and verified installation years were missing or the verified year occurred earlier or equal to the start of the study period (2009) has been excluded from analysis, yielding n=`r sfig8_n` segments. The graph shows that `r sfig8_0`% of the included segments had the correct installation year as per the city's data, and `r sfig8_pm1`% were accurate within a range of 1 year.

### Figure

```{r fig.width = 8, fig.height = 5}
sfig8$plot
```

### Files

```{r}

# Save pdf
ggsave(
	"../manuscript/figures/sfig-compare-calg.pdf",
	sfig8$plot,
	width = 18,
	height = 10
)

# Save png
ggsave(
	"../manuscript/figures/sfig-compare-calg.png",
	sfig8$plot,
	width = 18,
	height = 10
)
```

* [manuscript/figures/sfig-compare-calg.pdf](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/sfig-compare-calg.pdf)
* [manuscript/figures/sfig-compare-calg.png](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/sfig-compare-calg.png)

## Supplementary Figure 9: A comparative analysis between municipal data and verified data on the installation years for cycling infrastructure in Toronto, CA. {.tabset}

```{r}

# Create the plot
sfig9 <- plot_yearly_diff(
	toron,
	title = "Difference in Installation Years, Comparing City Data and Verified Data: Toronto, CA",
	out_data = TRUE
)

# Calc metrics for description
sfig9_n <- sum(sfig9$data$n)
sfig9_0 <- sfig9$data %>%  # perc correct
	filter(year_diff == 0) %>%
	pull(n_perc) %>%
	round(1)
sfig9_pm1 <- sfig9$data %>% # perc plus/minus 1
	filter(year_diff >= -1 & year_diff <= 1) %>%
	pull(n_perc) %>%
	sum %>%
	round(1)
```

Any data where a city provided and verified installation years were missing or the verified year occurred earlier or equal to the start of the study period (2009) has been excluded from analysis, yielding n=`r sfig9_n` segments. The graph shows that `r sfig9_0`% of the included segments had the correct installation year as per the city's data, and `r sfig9_pm1`% were accurate within a range of 1 year.

### Figure

```{r fig.width = 8, fig.height = 5}
sfig9$plot
```

### Files

```{r}

# Save pdf
ggsave(
	"../manuscript/figures/sfig-compare-toron.pdf",
	sfig9$plot,
	width = 18,
	height = 10
)

# Save png
ggsave(
	"../manuscript/figures/sfig-compare-toron.png",
	sfig9$plot,
	width = 18,
	height = 10
)
```

* [manuscript/figures/sfig-compare-toron.pdf](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/sfig-compare-toron.pdf)
* [manuscript/figures/sfig-compare-toron.png](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/sfig-compare-toron.png)

# Appendix 2 - Methodology

## Segment Inclusion Criteria for Vancouver {.tabset}

```{r}
sfig10 <- list()
sfig10$plot <- diag_criteria_details(criteria_data, "vancouver")
```

### Figure

```{r}
sfig10$plot
```

### Files

```{r}      

# Save pdf
sfig10$plot %>%
    export_svg %>%
    charToRaw %>%
    rsvg_pdf("../manuscript/figures/sfig-criteria-vanc.pdf")

# Save png
sfig10$plot %>%
    export_svg %>%
    charToRaw %>%
    rsvg_png("../manuscript/figures/sfig-criteria-vanc.png")
```

* [manuscript/figures/sfig-criteria-vanc.pdf](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/sfig-criteria-vanc.pdf)
* [manuscript/figures/sfig-criteria-vanc.png](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/sfig-criteria-vanc.png)

## Segment Inclusion Criteria for Calgary {.tabset}

```{r}
sfig11 <- list()
sfig11$plot <- diag_criteria_details(criteria_data, "calgary")
```

### Figure

```{r}
sfig11$plot
```

### Files

```{r}      

# Save pdf
sfig11$plot %>%
    export_svg %>%
    charToRaw %>%
    rsvg_pdf("../manuscript/figures/sfig-criteria-calg.pdf")

# Save png
sfig11$plot %>%
    export_svg %>%
    charToRaw %>%
    rsvg_png("../manuscript/figures/sfig-criteria-calg.png")
```

* [manuscript/figures/sfig-criteria-calg.pdf](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/sfig-criteria-calg.pdf)
* [manuscript/figures/sfig-criteria-calg.png](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/sfig-criteria-calg.png)

## Segment Inclusion Criteria for Toronto {.tabset}

```{r}
sfig12 <- list()
sfig12$plot <- diag_criteria_details(criteria_data, "toronto")
```

### Figure

```{r}
sfig12$plot
```

### Files

```{r}      

# Save pdf
sfig12$plot %>%
    export_svg %>%
    charToRaw %>%
    rsvg_pdf("../manuscript/figures/sfig-criteria-toron.pdf")

# Save png
sfig12$plot %>%
    export_svg %>%
    charToRaw %>%
    rsvg_png("../manuscript/figures/sfig-criteria-toron.png")
```

* [manuscript/figures/sfig-criteria-toron.pdf](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/sfig-criteria-toron.pdf)
* [manuscript/figures/sfig-criteria-toron.png](https://github.com/rrwen/recovr-infracycle/blob/main/manuscript/figures/sfig-criteria-toron.png)

# Contributions

Richard Wen developed reproducible R code and organized the data based on Konrad Samsel's draft manuscript and previous R code. Konrad Samsel prepared draft manuscript, raw data, and provided consultation on data and methods.

# Acknowledgements

Linda Rothman and Brice Batomen provided supervision, project administration, resources, funding, and review/editing for the draft manuscript.

# Software and Package Versions

R and R package versions:

```{r echo = FALSE}
sessionInfo()
```

RStudio version:

```{r echo = FALSE}
rstudioapi::versionInfo()
```